---
title: "Hydropower’s Next Chapter: Growth, Reservoirs, and Carbon"
description: "A standalone design-process blog post (EDS 240 FPM #4)"
author:
  - name: Lucian Blue Scher
    url: https://lucianbluescher.github.io
    affiliation: Master of Environmental Data Science Program @ The Bren School (UCSB)
    affiliation-url: https://ucsb-meds.github.io/
date: 2026-03-09
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
editor_options:
  chunk_output_type: inline
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

Hydropower is the world’s largest renewable electricity source, and in many grids it plays a uniquely valuable role: **it is dispatchable**. Operators can ramp generation up and down to match demand, complementing variable wind and solar. That flexibility is why many net-zero roadmaps still call for substantial hydropower growth, even as solar and wind dominate additions.

But hydropower is not automatically “low-carbon.” Dams create reservoirs, and reservoirs can emit greenhouse gases—especially methane—when flooded organic matter decomposes under low-oxygen conditions. Those emissions vary widely across geographies and designs, which means a hydropower buildout can look very different depending on **where** projects are built and **what kinds** of reservoirs they create.

This post is a standalone, story-style walk-through of selected final visuals from my infographic. The planning question I’m trying to make legible is:

- **If hydropower must expand, where are projects concentrated—and what environmental conditions are linked to higher carbon-intensity risk?**

## About the Data

::: panel-tabset

### IEA NZE (scenario)

The **IEA Net Zero by 2050 (NZE)** scenario provides the long-run context for my “hydro doubles, share declines” panel ([IEA report page](https://www.iea.org/reports/net-zero-by-2050)). I use the `NZE2021_AnnexA.csv` file (`data/NZE2021_AnnexA.csv`), filtering `Flow == "Electricity generation"` and `Unit == "TWh"` to isolate electricity production by source. From there, I focus on `Product` values `"Hydro"`, `"Solar PV"`, `"Wind"`, and `"Renewables"`, with `Year` and `Value` giving the time series in TWh. A `pivot_wider()` step turns this into a wide table where I compute “Other renewables” as a residual (`Renewables - Hydro - Solar PV - Wind`) and calculate hydropower’s percentage share of total renewables (`Hydro / Renewables * 100`) for annotation. I do not modify or re-scale the NZE scenario beyond this basic reshaping and arithmetic.

### FHReD (future dams)

The **Future Hydropower Reservoirs and Dams (FHReD)** database is my main source for planned hydropower projects (overview: [GlobalDamWatch FHReD](https://www.globaldamwatch.org/fhred/); associated publication: [Zarfl et al. (2015)](https://www.pnas.org/doi/10.1073/pnas.1509007112)). I work from the beta Excel file in this repo (`data/FHReD_2015_future_dams_Zarfl_et_al_beta_version/FHReD_2015_future_dams_Zarfl_et_al_beta_version.xlsx`, sheet 2), which lists proposed dams with attributes including basin identity, capacity, and coordinates. For this infographic I filter to rows with non-missing `Capacity (MW)`, `LAT_cleaned`, and `Lon_Cleaned`, and I construct a `basin` variable from `Major Basin` (falling back to `"Other basins"` when missing). I then summarize by `basin` to rank the top 10 “major basins” globally by planned MW (feeding both the palette and the treemap) and use the project coordinates as inputs to a spatial join against basin polygons for mapping.

### BasinATLAS (lev02 basins)

To move from project points to hydrologic regions, I use **BasinATLAS / HydroBASINS** polygons (see [HydroSHEDS HydroBASINS product page](https://www.hydrosheds.org/products/hydrobasins)). In this repo, the level-2 basin geometry is bundled as `data/basins_lev02_geom.rds`, which contains `hybas_id` and `MULTIPOLYGON` geometry in WGS84. I convert FHReD projects to an `sf` point layer using `LAT_cleaned` and `Lon_Cleaned`, repair basin geometries with `st_make_valid()`, and then run `st_join()` to attach each dam to the `hybas_id` that contains it. Summing `Capacity (MW)` by `hybas_id` yields a basin-level `future_mw` field that I map as a grey, square-root–scaled fill; the individual projects remain as colored points keyed to the major-basin palette.

### Li & He (biomes + CI patterns)

**Li & He (2022)** provides the carbon-intensity patterns I use as a conceptual overlay rather than as a full physical model ([paper DOI](https://doi.org/10.1016/j.rser.2022.112251)). Instead of ingesting raw reservoir time series, I hardcode published summary statistics from the supplementary tables into the script. From Table S2, I take median carbon intensities (kg CO₂e/MWh) for four broad biomes—Tundra, Boreal, Temperate, Tropical—and use these to define latitudinal “biome bands” on the map as contextual background. From Table S3, I extract median CI values for three depth classes and three trophic states at the **20-year GWP horizon** and build a 3×3 “scenario grid” index where each cell is `CI_depth_20y + CI_trophic_20y`. The grid is labeled as an **index constructed from marginal medians**, not an observed joint CI table; its purpose is to make the “highest-risk corner” (shallow + eutrophic) easy to communicate.

:::

## Set up
Here I load libraries and define the Okabe-Ito color palette and helper function for consistent basin coloring across panels. The palette is designed for high contrast and accessibility, and the helper function ensures that the same basin gets the same color in both the map and the treemap.
```{r}
#| label: setup
#| warning: false
#| message: false
#| code-fold: true
library(tidyverse)
library(here)
library(readxl)
library(scales)
library(sf)
library(ggnewscale)
library(ggrepel)
library(treemapify)
library(haven)
library(lubridate)
library(patchwork)

knitr::opts_chunk$set(dpi = 150)

# Okabe-Ito Extended: High contrast, maximum accessibility
# This is our single source of truth for basin colors across the map + treemap.
okabe_ito_11 <- c("#F0E442", # 1. Yangtze
                  "#009E73", # 2. Amazon
                  "#E69F00", # 3. Indus
                  "#CC79A7", # 4. Ganges - Brahmaputra
                  "#0072B2", # 5. Congo
                  "#117733", # 6. Mekong
                  "#D55E00", # 7. Salween
                  "#004E89", # 8. La Plata
                  "#56B4E9", # 9. Irrawady
                  "#442288", # 10. Nile
                  "#999999") # 11. Other basins (grey)

# Helper: take the 10 basin *names* (in rank order), return a named vector of colors.
# This ensures basin→color matching is stable across plots, even if data are filtered.
make_top10_basin_palette <- function(top10_basin_names) {
  if (length(top10_basin_names) != 10) {
    stop("make_top10_basin_palette() expects exactly 10 basin names.")
  }
  setNames(okabe_ito_11, c(paste0(1:10, ". ", top10_basin_names), "Other basins"))
}
```


## Top 10 Basins 
First I read and clean the FHReD dataset, rank basins by potential megawatts (MW), and build a palette for consistent coloring across the map and treemap. This chunk is the data wrangling backbone for our map and treemap.

```{r}
#| label: build-shared-fhred-and-palette
#| warning: false
#| message: false
#| code-fold: true
# FHReD (Future Hydropower Reservoirs and Dams) is the pipeline dataset:
# - each row is a proposed future hydropower project
# - we use this chunk to (1) read it, (2) clean key fields, (3) rank basins by total planned MW,
#   and (4) build a named palette so the same basin has the same color everywhere.
xl_path <- here("data/FHReD_2015_future_dams_Zarfl_et_al_beta_version/FHReD_2015_future_dams_Zarfl_et_al_beta_version.xlsx")
future <- read_xlsx(xl_path, sheet = 2)

future_geo <- future %>%
  # Only keep projects with the fields needed for mapping and aggregation
  filter(!is.na(`Capacity (MW)`), !is.na(LAT_cleaned), !is.na(Lon_Cleaned)) %>%
  mutate(
    # Basin name drives: top-10 ranking, labels, and color mapping
    basin = coalesce(`Major Basin`, "Other basins"),
    # FHReD contains a common spelling variant; standardize once so everything downstream matches.
    basin = stringr::str_replace_all(basin, "Ganges - Bramaputra", "Ganges - Brahmaputra")
  )

# Rank “major basins” by total planned capacity (MW).
# This is the ordering used for the top-10 labels and for basin colors.
basin_mw <- future_geo %>%
  group_by(basin) %>%
  summarise(basin_mw_total = sum(`Capacity (MW)`, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(basin_mw_total)) %>%
  mutate(rank = row_number())

# Top 10 basins by planned MW.
top_basins <- basin_mw %>% slice_head(n = 10) %>% pull(basin)

# Named color mapping used by all basin-colored layers (points on map + tiles in treemap).
basin_palette <- make_top10_basin_palette(top_basins)

future_geo <- future_geo %>%
  left_join(basin_mw, by = "basin") %>%
  mutate(
    # Create the ranked basin labels (e.g., “1. Yangtze”) used in both map + treemap.
    basin_ranked = if_else(
      basin %in% top_basins,
      paste0(match(basin, top_basins), ". ", basin),
      "Other basins"
    ),
    # Force a stable factor order so colors/labels don’t change between plots.
    basin_ranked = factor(basin_ranked, levels = names(basin_palette))
  )
```

## World Map

This map layers three different ideas on one geography. The background bands show broad **biome zones** (Li & He), basin polygons show **where total planned MW accumulates** (FHReD aggregated to BasinATLAS), and points show the locations and sizes of individual proposed dams. Together, it supports the thesis that future hydropower growth is concentrated in a handful of basins and latitudes where carbon-intensity risks may be higher.

```{r}
#| label: fig-map17
#| fig-width: 11
#| fig-height: 6
#| fig-cap: "Future hydropower on one map: biome bands (Li & He), basin polygons shaded by total future capacity (MW; FHReD aggregated to BasinATLAS lev02), and future dam projects as points colored by top-10 major basins."
#| fig-alt: "World map with pale biome latitude bands, gray basin polygons shaded by future hydropower capacity, and colored dam points; top-10 basin labels are repelled with leader lines."
#| warning: false
#| message: false
#| code-fold: true
# Li & He (2022) Table S2: median CI by biome (used as contextual background bands).
biome_ci <- tribble(
  ~biome,     ~ci_median,
  "Tundra",      43.8,
  "Boreal",     104.4,
  "Temperate",   54.8,
  "Tropical",    91.0
)

# Helper to assign a simple biome label from latitude (a coarse approximation used only for bands).
assign_biome <- function(lat) {
  case_when(
    lat >= 60 | lat <= -60 ~ "Tundra",
    (lat >= 50 & lat < 60) | (lat <= -50 & lat > -60) ~ "Boreal",
    (lat >= 23 & lat < 50) | (lat <= -23 & lat > -50) ~ "Temperate",
    lat > -23 & lat < 23 ~ "Tropical",
    TRUE ~ NA_character_
  )
}

# Attach biome labels to FHReD points so we can summarize and label consistently.
future_geo_map <- future_geo %>%
  mutate(biome = assign_biome(LAT_cleaned)) %>%
  left_join(biome_ci, by = "biome")

# Latitude-band rectangles for biome zones (background layer).
lat_bands <- tribble(
  ~ymin, ~ymax, ~biome,
  -90,   -60,   "Tundra",
  -60,   -50,   "Boreal",
  -50,   -23,   "Temperate",
  -23,    23,   "Tropical",
   23,    50,   "Temperate",
   50,    60,   "Boreal",
   60,    90,   "Tundra"
) %>%
  left_join(biome_ci, by = "biome")

# Biome band colors (background). These are separate from the basin palette.
pal_ci <- c(
  "Tundra"    = "#D8E9D9",
  "Boreal"    = "#E0E0E0",
  "Temperate" = "#A8D1AA", 
  "Tropical"  = "#58A86F" 
)

# Basin polygons (BasinATLAS lev02). These are used to spatially aggregate FHReD points.
basins_lev02 <- readRDS(here("data/basins_lev02_geom.rds"))
basins_lev02 <- sf::st_as_sf(basins_lev02)
if (is.na(sf::st_crs(basins_lev02))) sf::st_crs(basins_lev02) <- 4326

# Some global basin polygons can contain invalid rings; fix geometries before spatial join.
sf::sf_use_s2(FALSE)
basins_lev02 <- sf::st_make_valid(basins_lev02)
basins_lev02 <- sf::st_collection_extract(basins_lev02, "POLYGON")
basins_lev02 <- sf::st_cast(basins_lev02, "MULTIPOLYGON", warn = FALSE)

# Convert FHReD projects to points for spatial join.
future_pts <- future_geo_map %>%
  st_as_sf(coords = c("Lon_Cleaned", "LAT_cleaned"), crs = 4326, remove = FALSE)

# Spatial join: attach BasinATLAS `hybas_id` to each FHReD project.
future_with_basin <- st_join(
  future_pts,
  basins_lev02 %>% select(hybas_id),
  left = FALSE
)
sf::sf_use_s2(TRUE)

# Aggregate planned MW (from FHReD) into BasinATLAS polygons.
basin_stats <- future_with_basin %>%
  st_drop_geometry() %>%
  group_by(hybas_id) %>%
  summarise(
    future_mw = sum(`Capacity (MW)`, na.rm = TRUE),
    .groups = "drop"
  )

basin_map <- basins_lev02 %>%
  left_join(basin_stats, by = "hybas_id") %>%
  mutate(future_mw = replace_na(future_mw, 0))

# Label anchors: take the median project coordinate per top-10 basin.
# Nudges are manual so labels can be pushed over open ocean for readability.
basin_labels <- future_geo_map %>%
  filter(basin %in% top_basins) %>%
  group_by(basin) %>%
  summarise(
    anchor_x = median(Lon_Cleaned, na.rm = TRUE),
    anchor_y = median(LAT_cleaned, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    basin_mw %>%
      filter(rank <= 10) %>%
      transmute(
        basin,
        label_rank = paste0(rank, ". ", basin)
      ),
    by = "basin"
  ) %>%
  mutate(
    label = label_rank,
    nudge_x = case_when( # + for east, - for west
      basin == "Yangtze"  ~ 60,
      basin == "Mekong"   ~ 40,
      basin == "Amazon"   ~ 30,
      basin == "La Plata" ~ 25,
      basin == "Congo"    ~ -30,
      basin == "Nile"     ~ 20,
      basin == "Irrawaddy" ~ -15,
      basin == "Salween"  ~ 40,
      basin == "Indus"    ~ -10,
      basin == "Ganges - Brahmaputra"   ~ 10,
      TRUE ~ 0
    ),
    nudge_y = case_when( # + for north, - for south
      basin == "Yangtze"  ~ 10,
      basin == "Mekong"   ~ -15,
      basin == "Amazon"   ~ 10,
      basin == "La Plata" ~ -5,
      basin == "Congo"    ~ -15,
      basin == "Nile"     ~ -5,
      basin == "Irrawaddy" ~ -25,
      basin == "Salween"  ~ -5,
      basin == "Indus"    ~ 10,
      basin == "Ganges - Brahmaputra"   ~ 25,
      TRUE ~ 0
    )
  )

p_map <- ggplot() +
  geom_rect(
    data = lat_bands,
    aes(xmin = -180, xmax = 180, ymin = ymin, ymax = ymax, fill = biome),
    color = NA,
    alpha = 0.35
  ) +
  scale_fill_manual(values = pal_ci, name = "Biome") +
  ggnewscale::new_scale_fill() +
  geom_sf(
    data = basin_map,
    aes(fill = future_mw),
    color = NA,
    inherit.aes = FALSE
  ) +
  scale_fill_gradient(
    low = "grey95",
    high = "grey20",
    trans = "sqrt",
    labels = comma,
    name = "Basin capacity (MW)"
  ) +
  geom_point(
    data = future_geo_map,
    aes(
      x = Lon_Cleaned,
      y = LAT_cleaned,
      size = `Capacity (MW)`,
      color = basin_ranked
    ),
    shape = 16,
    alpha = 0.85,
    inherit.aes = FALSE
  ) +
  scale_size_continuous(range = c(1.2, 9), labels = comma, name = "Dam capacity (MW)") +
  scale_color_manual(values = basin_palette, guide = "none") +
  ggrepel::geom_text_repel(
    data = basin_labels,
    aes(x = anchor_x, y = anchor_y, label = label),
    color = "#000000",
    size = 3,
    fontface = "bold",
    family = "Times New Roman",
    lineheight = 0.9,
    seed = 123,
    nudge_x = basin_labels$nudge_x,
    nudge_y = basin_labels$nudge_y,
    box.padding = 0.5,
    point.padding = 0.5,
    segment.color = "#000000",
    segment.size = 0.4,
    segment.alpha = 0.95,
    segment.curvature = 0.2,
    segment.angle = 20,
    segment.ncp = 3,
    min.segment.length = 0,
    max.overlaps = 20,
    inherit.aes = FALSE
  ) +
  coord_sf(xlim = c(-180, 180), ylim = c(-60, 75)) +
  labs(
    # title = "Future hydropower clusters in top 10 future MW producing basins",
    # subtitle = "Basins shaded by total planned capacity (MW). Dams colored by top-10 major basins; other basins are grey.",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.text = element_text(family = "Times New Roman"),
    legend.title = element_text(face = "bold", family = "Times New Roman")
  )

p_map

# Save for layout work (non-transparent background for the map).
ggsave(
  filename = here("images", "map_future_basins_biomes.pdf"),
  plot = p_map,
  device = cairo_pdf,
  width = 11,
  height = 6,
  units = "in",
  bg = "white"
)
```

## Treemap

This treemap collapses the FHReD pipeline into one simple message: a few basins dominate the future buildout. Tile area encodes planned MW, and the ranked basin labels match the map so the same basin has the same name and color across panels.

```{r}
#| label: fig-treemap14
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Future hydropower capacity by major river basin (FHReD), showing the top 10 basins plus all others aggregated."
#| fig-alt: "Treemap where each rectangle area represents planned hydropower megawatts by river basin; top basins are colored distinctly and labeled with megawatts."
#| warning: false
#| message: false
#| code-fold: true

# Treemap (tile area = MW; colors = basin palette)
treemap_data <- future_geo %>%
  group_by(basin_ranked) %>%
  summarise(mw = sum(`Capacity (MW)`, na.rm = TRUE), .groups = "drop") %>%
  filter(mw > 0) %>%
  mutate(
    # Split labels so the basin name can scale independently of the MW line
    basin_name = as.character(basin_ranked),
    mw_label   = paste0("(", scales::comma(round(mw)), " MW)")
  )

p_treemap <- ggplot(treemap_data, aes(area = mw, fill = basin_ranked)) +
  geom_treemap(layout = "squarified", color = "white", linewidth = 0.5) +
  # Big basin name: scales to fill each rectangle
  geom_treemap_text(
    aes(label = basin_name),
    place = "centre",
    grow = TRUE,
    family = "Times New Roman",
   # fontface = "bold",
    color = "black",
    min.size = 1,
    padding.y = grid::unit(2, "mm")
  ) +
  # Small MW line: fixed size near bottom
  geom_treemap_text(
    aes(label = mw_label),
    place = "bottom",
    size = 10,
    family = "Times New Roman",
    fontface = "plain",
    color = "black",
    min.size = 1,
    padding.y = grid::unit(2, "mm")
  ) +
  scale_fill_manual(values = basin_palette, guide = "none") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )

p_treemap

# Save with transparent background for layout.
ggsave(
  filename = here("images", "treemap_future_capacity_by_basin.pdf"),
  plot = p_treemap,
  device = cairo_pdf,
  width = 8,
  height = 5,
  units = "in",
  bg = "transparent"
)
```

## “Hydropower sun”

This “sun” plot is a compact way to show that hydropower is not just “how much,” but also **when**—it has strong daily and seasonal patterns. While this panel is a regional example, it supports the story’s framing that hydropower’s value is partly its timing and dispatchability.

```{r}
#| label: fig-sun01
#| fig-width: 6
#| fig-height: 6
#| fig-cap: "Hydropower “sun”: months at the core, hours as corona (regional example). Brighter = more generation."
#| fig-alt: "Circular heatmap with months in the inner ring and hourly rays in the outer ring, colored from pale to dark orange to show higher hydropower generation."
#| warning: false
#| message: false
#| code-fold: true
dta <- read_dta(here::here("data/marginal-generation-offsets-data.dta"))

inner_radius <- 1.5

inner_month <- dta %>%
  mutate(
    month_num = month(dt),
    month_lab = factor(month.abb[month_num], levels = month.abb)
  ) %>%
  group_by(month_lab) %>%
  summarise(h_mean = mean(h, na.rm = TRUE), .groups = "drop") %>%
  mutate(h_norm = h_mean / max(h_mean, na.rm = TRUE))

month_slots <- tibble(
  month_lab = factor(month.abb, levels = month.abb),
  x_center  = seq(1, 23, by = 2)
)

inner_month_long <- inner_month %>% inner_join(month_slots, by = "month_lab")

outer_hour <- dta %>%
  mutate(hour = hour(dt)) %>%
  group_by(hour) %>%
  summarise(h_mean = mean(h, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    h_norm   = h_mean / max(h_mean, na.rm = TRUE),
    hour_rot = (hour - 12) %% 24,
    inner_r  = inner_radius + 0.1,
    outer_r  = inner_r + 0.9 * h_norm
  )

label_r <- inner_radius + 1.15
time_labels <- tibble(
  hour_rot = c(0, 6, 12, 18),
  label = c("12pm", "6pm", "12am", "6am"),
  hjust = c(0.5, -0.1, 0.5, 1.1),
  vjust = c(-0.5, 0.5, 1.2, 0.5)
)

month_labels <- tibble(
  x_center = c(1, 7, 13, 19),
  y_pos = inner_radius * 0.6,
  label = c("Jan", "Apr", "Jul", "Oct")
)

p_sun <- ggplot() +
  geom_col(
    data = inner_month_long,
    aes(x = x_center, y = inner_radius, fill = h_norm),
    width = 2, color = NA
  ) +
  geom_segment(
    data = outer_hour,
    aes(x = hour_rot, xend = hour_rot, y = inner_r, yend = outer_r, color = h_norm),
    lineend = "round", linewidth = 1.0, alpha = 0.9
  ) +
  geom_text(
    data = time_labels,
    aes(x = hour_rot, y = label_r, label = label),
    hjust = time_labels$hjust,
    vjust = time_labels$vjust,
    size = 2.8,
    color = "black",
    family = "Times New Roman"
  ) +
  geom_text(
    data = month_labels,
    aes(x = x_center, y = y_pos, label = label),
    size = 2.5,
    color = "black",
    family = "Times New Roman"
  ) +
  coord_polar(theta = "x", start = 0) +
  scale_colour_gradientn(
    colours = c("#fff7bc", "#fee391", "#fec44f", "#fe9929", "#d95f0e", "#993404"),
    name = "Avg hydropower\n(normalized)"
  ) +
  scale_fill_gradientn(
    colours = c("#fff7bc", "#fee391", "#fec44f", "#fe9929", "#d95f0e", "#993404"),
    guide = "none"
  ) +
  scale_y_continuous(limits = c(0, inner_radius + 1.3), breaks = NULL) +
  labs(
    # title = "Hydropower “sun”: months at the core, hours as rays",
    # subtitle = "12:00 at top | Brighter = more generation",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.text = element_text(family = "Times New Roman"),
    legend.title = element_text(face = "bold", family = "Times New Roman"),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )

p_sun

# Save with transparent background for layout.
ggsave(
  filename = here("images", "sun_hydropower_timing.pdf"),
  plot = p_sun,
  device = cairo_pdf,
  width = 6,
  height = 6,
  units = "in",
  bg = "transparent"
)
```

## Stacked Area Chart

This panel uses the IEA NZE scenario to show why hydropower remains in the net-zero conversation: it grows in absolute terms, even as solar and wind grow faster. The vertical reference line marks “now-ish” to separate recent history from the scenario period.

```{r}
#| label: fig-renew03
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "In the IEA NZE scenario, hydropower generation increases, but solar and wind grow faster—so hydropower’s share of renewable electricity declines."
#| fig-alt: "Stacked area chart of renewable electricity generation by source from 2019 to 2050; hydropower grows but is a smaller share by 2050 as solar and wind expand."
#| warning: false
#| message: false
#| code-fold: true
nze <- read_csv(here::here("data/NZE2021_AnnexA.csv"), show_col_types = FALSE)

generation <- nze %>%
  filter(
    Flow == "Electricity generation",
    Unit == "TWh",
    Product %in% c("Hydro", "Solar PV", "Wind", "Renewables")
  ) %>%
  select(Product, Year, Value) %>%
  pivot_wider(names_from = Product, values_from = Value) %>%
  mutate(
    Other_Renewables = Renewables - Hydro - `Solar PV` - Wind,
    Hydro_Share_Renewables = (Hydro / Renewables) * 100
  )

viz_data <- generation %>%
  select(Year, Hydropower = Hydro, `Solar PV`, Wind, Other_Renewables) %>%
  pivot_longer(cols = -Year, names_to = "Source", values_to = "Generation") %>%
  mutate(
    Source = factor(Source, levels = c("Other_Renewables", "Wind", "Solar PV", "Hydropower")),
    Source_Label = case_when(
      Source == "Other_Renewables" ~ "Other",
      Source == "Solar PV" ~ "Solar PV",
      Source == "Wind" ~ "Wind",
      TRUE ~ "Hydropower"
    )
  )

p_mix <- ggplot(viz_data, aes(x = Year, y = Generation, fill = Source_Label)) +
  geom_area(alpha = 0.85, color = "#ffffff", linewidth = 0.3) +
  # vertical dotted line at 2026
  geom_vline(
    xintercept = 2026,
    linetype   = "dotted",
    color      = "gray40",
    linewidth  = 0.5
  ) +
  scale_fill_manual(
    values = c("Hydropower" = "#1c5d99", "Solar PV" = "#639fab",
               "Wind" = "#bbcde5", "Other" = "#222222"),
    breaks = c("Hydropower", "Solar PV", "Wind", "Other"),
    guide = guide_legend(reverse = TRUE)
  ) +
  scale_x_continuous(breaks = c(2019, 2026, 2030, 2040, 2050), expand = expansion(mult = c(0, 0.08))) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(x = "Year", y = "Electricity (TWh)", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position      = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.direction     = "vertical",
    legend.box           = "vertical",
    legend.background    = element_rect(fill = scales::alpha("white", 0.75), color = NA),
    legend.key           = element_rect(fill = NA, color = NA),
    legend.text          = element_text(family = "Times New Roman"),
    axis.text            = element_text(family = "Times New Roman"),
    panel.grid.minor     = element_blank(),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )
p_mix

# Save with transparent background for layout.
ggsave(
  filename = here("images", "stacked_area_nze_renewables_mix.pdf"),
  plot = p_mix,
  device = cairo_pdf,
  width = 6,
  height = 5,
  units = "in",
  bg = "transparent"
)
```

## Heatmap

This 3×3 scenario matrix (heatmap) uses Li & He’s reported medians to illustrate how carbon-intensity risk changes across depth and trophic conditions. It’s an index (sum of marginal medians) meant for communication and comparison, not a modeled joint prediction for any specific dam.

```{r}
#| label: fig-ci18
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Scenario grid (20-year horizon): an index built from Li & He (2022) marginal medians by depth and trophic state. Higher values indicate higher carbon-intensity risk conditions."
#| fig-alt: "3x3 heatmap grid with trophic state on x-axis and reservoir depth on y-axis; darker colors and larger numbers indicate higher index carbon intensity."
#| warning: false
#| message: false
#| code-fold: true
trophic_data <- tibble(
  Category = factor(c("Oligotrophic", "Mesotrophic", "Eutrophic"),
                    levels = c("Oligotrophic", "Mesotrophic", "Eutrophic")),
  CI_20yr  = c(67.2, 116.0, 248.0),
  CI_100yr = c(55.6, 69.1, 116.7)
)

depth_data <- tibble(
  Category = factor(c("Shallow (<20 m)", "Medium (20–100 m)", "Deep (>100 m)"),
                    levels = c("Deep (>100 m)", "Medium (20–100 m)", "Shallow (<20 m)")),
  CI_20yr  = c(233.0, 114.3, 35.6),
  CI_100yr = c(124.5, 66.0, 27.7)
)

trophic_20 <- trophic_data %>%
  transmute(trophic = as.character(Category), ci_trophic_20 = CI_20yr)

depth_20 <- depth_data %>%
  transmute(depth = as.character(Category), ci_depth_20 = CI_20yr)

scenario_grid <- tidyr::expand_grid(depth_20, trophic_20) %>%
  mutate(
    ci_index_20 = ci_depth_20 + ci_trophic_20,
    # factor with custom labels for y-axis
    depth = factor(
      depth,
      levels = c("Deep (>100 m)", "Medium (20–100 m)", "Shallow (<20 m)"),
      labels = c("Deep\n(>100 m)", "Medium\n(20–100 m)", "Shallow\n(<20 m)")
    ),
    # factor with custom labels for x-axis
    trophic = factor(
      trophic,
      levels = c("Oligotrophic", "Mesotrophic", "Eutrophic"),
      labels = c("Oligotrophic\n(low nutrients)",
                 "Mesotrophic",
                 "Eutrophic\n(high nutrients)")
    )
  )

p_grid <- ggplot(scenario_grid, aes(x = trophic, y = depth, fill = ci_index_20)) +
  geom_tile(color = "white", linewidth = 0.7) +
  geom_text(
    aes(label = round(ci_index_20)),
    color   = "white",
    fontface = "bold",
    size    = 6,
    family  = "Times New Roman"
  ) +
  scale_fill_viridis_c(
    option    = "magma",
    direction = 1,
    begin     = 0.1,
    end       = 0.9,
    name      = "Index CI\n(kg CO2e/MWh)"
  ) +
  labs(
    x = "Trophic state",
    y = "Reservoir depth"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid       = element_blank(),
    text             = element_text(family = "Times New Roman"),  # global
    axis.text        = element_text(family = "Times New Roman", color = "black"),
    axis.title       = element_text(family = "Times New Roman", face = "bold"),
    legend.title     = element_text(family = "Times New Roman", face = "bold"),
    legend.text      = element_text(family = "Times New Roman"),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )

p_grid

# Save with transparent background for layout.
ggsave(
  filename = here("images", "heatmap_lihe_scenario_grid.pdf"),
  plot = p_grid,
  device = cairo_pdf,
  width = 8,
  height = 5,
  units = "in",
  bg = "transparent"
)
```

## Design Choices

::: panel-tabset

### Graphic form

- **Macro → micro storytelling**: I start with the energy-system framing (renewables mix), then move into *where* future projects cluster (map + treemap), and finally into a simple *risk-conditions* matrix (scenario grid).
- **Aggregation choice**: I use basins because hydrologic impacts (flows, sediment, habitat connectivity) follow rivers, not borders.

### Text

- **Titles as takeaways**: Each plot title is written as a claim rather than a chart description.
- **Subtitles for decoding**: Subtitles carry the mapping rules (what color/size means) to keep the body text readable.

### Themes

- **Minimal scaffolding**: I remove gridlines where they don’t add value so the data shapes carry the narrative.
- **Consistent hierarchy**: Bold titles, lighter subtitles, and smaller captions so readers know where to look first.

### Colors

- **One categorical system reused**: Top-10 basin colors come from a single named palette so a basin keeps the same identity across plots.
- **Color restraint**: Basin polygons are grayscale (continuous MW) so basin-identity colors on dam points can pop.

### Typography

- **Functional legibility**: Label sizes are chosen for “poster distance.”
- **Map labels**: Repelled text and leader lines reduce overlap, especially in Southeast Asia.

### General design

- **Reduce cognitive load**: Each panel uses one dominant mapping (area, color, size) and avoids unnecessary overlays.
- **Avoid redundancy**: I removed dotted-line overlays in the renewables plot to keep the message focused.

### Contextualizing data

- **State limitations**: FHReD is a pipeline database; the scenario grid is an index built from marginal medians, not observed joint CI.
- **Show units**: MW and TWh are included directly in labels/captions.

### Centering the primary message

- **Decision points**: Hydropower expansion isn’t “good” or “bad” in isolation; impacts depend on siting and conditions.
- **Concentration matters**: Highlighting “a few basins” makes the governance and targeting problem feel tractable.

### Accessibility

- **Colorblind-safe categorical palette**: Okabe-Ito provides high-contrast, accessible colors for basin identity.
- **Alt text**: Each figure includes alt text describing encodings and structure.

### DEI lens

- **Avoid inevitability framing**: I emphasize choices and trade-offs rather than “the world will build these dams.”
- **Place-based impacts**: Basin framing highlights downstream communities and ecosystems that often cross borders.

:::


