---
title: "Hydropower Carbon Intensity — Li & He (2022) Summaries"
author: "Your Name"
format:
  html:
    toc: true
    toc-depth: 2
  pdf:
    documentclass: article
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center"
)
library(tidyverse)
library(scales)
```

# Data from Li & He (2022)

All numbers below are taken from **Li & He (2022), _Renewable and Sustainable Energy Reviews 162, 112433_**, and its supplementary Tables S2–S3 and Figure S4 (which itself cites IPCC 2013 for non‑hydro energy intensities).

```{r lihe-data}
# 1. Energy-type carbon intensities (medians) from Li & He text (IPCC 2013 values)
energy_ci <- tribble(
  ~source,         ~ci_median,
  "Coal",          1000,  # kg CO2e/MWh
  "Oil",           780,
  "Natural gas",   490,
  "Hydropower",     63,   # median CI for existing hydro in Li & He
  "Solar PV",       48,
  "Wind",           12,
  "Nuclear",        12
)

# 2. Biome-level GHG flux & CI (Table S2, 100-year horizon)
biome_ci <- tribble(
  ~biome,     ~flux_min, ~flux_med, ~flux_max, ~ci_median,
  "Tundra",     489.8,     582.6,    1195.4,     43.8,
  "Boreal",     337.2,     788.9,    1230.9,    104.4,
  "Temperate",  468.7,    1046.9,    1568.7,     54.8,
  "Tropical",   855.3,    1955.5,    2323.1,     91.0
)

# 3. Trophic-level GHG components & CI (Table S3, 100-year horizon)
trophic_ci <- tribble(
  ~trophic,        ~co2,  ~ch4_diff, ~ch4_ebu, ~n2o,  ~ci_median,
  "Oligotrophic",  0.009, 0.002,     0.010,    0.002,  55.6,
  "Mesotrophic",   0.036, 0.019,     0.070,    0.009,  69.1,
  "Eutrophic",     0.029, 0.044,     0.140,    0.010, 116.7
)

# 4. Depth-class GHG components & CI (Table S3, 100-year horizon)
depth_ci <- tribble(
  ~depth_class, ~co2,  ~ch4_diff, ~ch4_ebu, ~n2o,  ~ci_median,
  "Shallow (<20 m)", 0.049, 0.043, 0.156, 0.016, 124.5,
  "Medium (20–100 m)", 0.024, 0.022, 0.066, 0.005, 66.0,
  "Deep (>100 m)", 0.0005, 0.0006, 0.0016, 0.00009, 27.7
)

# 5. Global totals by pathway (Table S3, "Total" row, 100-year horizon)
pathways_total <- tribble(
  ~gas,   ~component,      ~pg_co2eq_yr,
  "CO2",  "CO2",             0.074,
  "CH4",  "CH4 diffusion",   0.066,
  "CH4",  "CH4 ebullition",  0.220,
  "N2O",  "N2O",             0.021
) %>%
  mutate(
    total = sum(pg_co2eq_yr),
    share = pg_co2eq_yr / total * 100
  )

# 6. Existing vs existing+future hydropower CI (Table 2, "All" row)
ci_existing_future <- tribble(
  ~group,             ~ghg_pg, ~ci_median, ~pct_above_80,
  "Existing dams",      0.26,     63.0,       44.0,  # % failing 80 kg target (text)
  "Future dams",        0.11,       NA,       66.4,  # % future >80
  "Existing + future",  0.37,    131.5,         NA
)
```

---

# Graph 1 — Energy-type carbon intensity comparison

```{r g1-energy-ci}
ggplot(energy_ci, aes(x = fct_reorder(source, ci_median), y = ci_median,
                      fill = source == "Hydropower")) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#1b9e77", "FALSE" = "grey75"), guide = "none") +
  labs(
    title = "Life-cycle carbon intensity across power technologies",
    subtitle = "Medians from IPCC (2013) as reported in Li & He (2022)",
    x = NULL,
    y = "kg CO\u2082e per MWh"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

**Takeaway:** Hydropower is far cleaner than fossil fuels but generally dirtier than wind, solar, and nuclear (Li & He 2022; IPCC 2013).

---

# Graph 2 — Biome carbon intensity (Li & He Table S2)

```{r g2-biome-ci}
ggplot(biome_ci, aes(x = biome, y = ci_median)) +
  geom_col(fill = "#3182bd", width = 0.7) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "red") +
  annotate("text", x = 4.2, y = 82, label = "80 kg CO\u2082e/MWh (2030 Agenda target)",
           hjust = 1, size = 3, color = "red") +
  labs(
    title = "Median hydropower carbon intensity by biome",
    subtitle = "Li & He (2022), Table S2 — 100-year GWP horizon",
    x = NULL,
    y = "kg CO\u2082e per MWh"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

**Takeaway:** Boreal and tropical hydropower tend to have higher carbon intensity; many temperate and tundra reservoirs sit below the 80 kg CO₂e/MWh “low‑carbon” line.

---

# Graph 3 — Trophic level vs carbon intensity (Table S3)

```{r g3-trophic-ci}
ggplot(trophic_ci, aes(x = trophic, y = ci_median)) +
  geom_col(fill = "#fb9a99", width = 0.7) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "red") +
  labs(
    title = "Reservoir carbon intensity by trophic level",
    subtitle = "Li & He (2022), Table S3 — 100-year horizon",
    x = NULL,
    y = "kg CO\u2082e per MWh"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

**Takeaway:** Eutrophic reservoirs (nutrient‑rich, high Chla) have much higher CI than oligotrophic and mesotrophic systems.

---

# Graph 4 — Depth class vs carbon intensity (Table S3)

```{r g4-depth-ci}
ggplot(depth_ci, aes(x = depth_class, y = ci_median)) +
  geom_col(fill = "#a6cee3", width = 0.7) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "red") +
  labs(
    title = "Reservoir carbon intensity by depth class",
    subtitle = "Li & He (2022), Table S3 — 100-year horizon",
    x = NULL,
    y = "kg CO\u2082e per MWh"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

**Takeaway:** Shallow reservoirs (<20 m) are the main problem: their median CI is well above the low‑carbon threshold, while deep reservoirs are comfortably below it.

---

# Graph 5 — Global GHG by pathway (CO₂, CH₄, N₂O)

```{r g5-pathways}
pathways_total %>%
  mutate(component = fct_reorder(component, pg_co2eq_yr)) %>%
  ggplot(aes(x = component, y = pg_co2eq_yr, fill = gas)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("CO2" = "#3182bd", "CH4" = "#e6550d", "N2O" = "#31a354")) +
  labs(
    title = "Global reservoir GHG emissions by pathway (100-year horizon)",
    subtitle = "Li & He (2022), Table S3 — Total row",
    x = NULL,
    y = "Pg CO\u2082e per year",
    fill = "Gas"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

**Takeaway:** CH₄ (especially ebullitive CH₄) dominates global reservoir warming; CO₂ and N₂O are secondary contributors.

---

# Graph 6 — Pathway shares (stacked bar)

```{r g6-pathway-share}
ggplot(pathways_total, aes(x = "Reservoir GHG", y = share, fill = component)) +
  geom_col(width = 0.4, color = "white") +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  labs(
    title = "Share of global reservoir warming by pathway",
    subtitle = "Li & He (2022) — CH\u2084 paths supply ~72.5% of total",
    x = NULL,
    y = "% of total Pg CO\u2082e/yr",
    fill = "Pathway"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid.minor = element_blank())
```

**Takeaway:** Ebullitive + diffusive CH₄ account for roughly three‑quarters of reservoir GHG impact, reinforcing the importance of methane pathways.

---

# Graph 7 — Existing vs future hydropower: CI & 80 kg threshold

```{r g7-existing-future}
ci_plot <- ci_existing_future %>%
  filter(group %in% c("Existing dams", "Existing + future"))

ggplot(ci_plot, aes(x = group, y = ci_median)) +
  geom_col(fill = "#9ecae1", width = 0.6) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "red") +
  labs(
    title = "Median hydropower carbon intensity: today vs with planned dams",
    subtitle = "Li & He (2022), Table 2 — 100-year horizon",
    x = NULL,
    y = "kg CO\u2082e per MWh"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

**Takeaway:** Adding planned/under‑construction dams increases the global median CI from ~63 to ~132 kg CO₂e/MWh, moving hydropower further from the “low‑carbon” target.

---

# Graph 8 — Share of reservoirs failing the 80 kg target

```{r g8-share-failing}
share_failing <- ci_existing_future %>%
  filter(group %in% c("Existing dams", "Future dams")) %>%
  mutate(
    pct_above_80 = pct_above_80,
    pct_below_80 = 100 - pct_above_80
  ) %>%
  pivot_longer(cols = starts_with("pct_"),
               names_to = "category", values_to = "percent") %>%
  mutate(
    category = recode(category,
      "pct_above_80" = ">\n80 kg CO\u2082e/MWh",
      "pct_below_80" = "\u2264\n80 kg CO\u2082e/MWh"
    )
  )

ggplot(share_failing, aes(x = group, y = percent, fill = category)) +
  geom_col(width = 0.6, color = "white") +
  scale_fill_manual(values = c(">\n80 kg CO\u2082e/MWh" = "#e34a33",
                               "\u2264\n80 kg CO\u2082e/MWh" = "#31a354")) +
  labs(
    title = "Share of reservoirs meeting the 80 kg CO\u2082e/MWh '2030 Agenda' goal",
    subtitle = "Li & He (2022, main text) — existing vs future hydropower dams",
    x = NULL,
    y = "% of reservoirs",
    fill = NULL
  ) +
  coord_flip() +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

**Takeaway:** Only about 56% of existing dams and 34% of planned dams meet the 80 kg CO₂e/MWh goal; most future hydropower capacity lies in high‑CI reservoirs.

---

# Citation

> Li, M., & He, N. (2022). Carbon intensity of global existing and future hydropower reservoirs. *Renewable and Sustainable Energy Reviews, 162*, 112433.


