---
title: "Life Cycle Emissions Explorations: Hydropower's Climate Profile"
author: "Your Name"
date: today
format:
  pdf:
    documentclass: article
    fig-height: 6
    fig-width: 10
    keep-tex: false
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Introduction

This analysis explores life cycle greenhouse gas emissions data from the GLEAM/NREL Life Cycle Assessment Harmonization Project. We focus on hydropower's emissions profile compared to other electricity generation technologies across major categories, highlighting hydropower's high variability and its position between fossil fuels and other low-carbon options.

# Load Packages

```{r load-packages}
library(tidyverse)
library(ggplot2)
library(readxl)
library(janitor)
library(scales)
library(patchwork)
library(ggrepel)
library(treemapify)
```

# Import Data

```{r import-data}
# GLEAM LCA data: Category, Technology, Upstream, Ongoing Non-Combustion, Total Life Cycle (each with min, 1q, med, 3q, max)

gleam_file <- "data/EF_Table_FINAL.xlsx"
if (!file.exists(gleam_file)) stop("EF_Table_FINAL.xlsx not found at data/EF_Table_FINAL.xlsx")

gleam_raw <- read_excel(gleam_file, skip = 2, .name_repair = "minimal")
gleam_lca <- clean_names(gleam_raw)

n_cols <- ncol(gleam_lca)
gleam_lca <- gleam_lca %>%
  select(1, 2, 3:27, (n_cols-3):n_cols) %>%
  rename(
    category = x,
    generation_technology = x_2,
    upstream_min = min, upstream_1q = x1q, upstream_med = med, upstream_3q = x3q, upstream_max = max,
    combustion_min = min_2, combustion_1q = x1q_2, combustion_med = med_2, combustion_3q = x3q_2, combustion_max = max_2,
    non_combustion_min = min_3, non_combustion_1q = x1q_3, non_combustion_med = med_3, non_combustion_3q = x3q_3, non_combustion_max = max_3,
    downstream_min = min_4, downstream_1q = x1q_4, downstream_med = med_4, downstream_3q = x3q_4, downstream_max = max_4,
    total_min = min_5, total_1q = x1q_5, total_med = med_5, total_3q = x3q_5, total_max = max_5
  )

numeric_cols <- c("upstream_min", "upstream_1q", "upstream_med", "upstream_3q", "upstream_max",
                  "combustion_min", "combustion_1q", "combustion_med", "combustion_3q", "combustion_max",
                  "non_combustion_min", "non_combustion_1q", "non_combustion_med", "non_combustion_3q", "non_combustion_max",
                  "downstream_min", "downstream_1q", "downstream_med", "downstream_3q", "downstream_max",
                  "total_min", "total_1q", "total_med", "total_3q", "total_max")

gleam_lca <- gleam_lca %>%
  mutate(across(all_of(numeric_cols), ~ {
    val <- as.character(.x)
    case_when(
      is.na(val) | val == "NA" | val == "NR" | val == "--" ~ NA_real_,
      str_detect(val, "^<") ~ as.numeric(str_extract(val, "\\d+")) * 0.5,
      TRUE ~ suppressWarnings(as.numeric(val))
    )
  })) %>%
  filter(!is.na(generation_technology),
         generation_technology != "Generation Technology",
         !str_detect(generation_technology, "^NA$"),
         !is.na(total_med) | !is.na(total_min) | !is.na(total_max))

gleam_lca <- gleam_lca %>%
  mutate(is_hydropower = str_detect(generation_technology, "Hydropower"))
```

# Data Wrangling

Filter to seven key technologies and assign display names and type (Hydropower, Fossil, Other) for coloring.

```{r wrangle-data}
gleam_clean <- gleam_lca %>%
  filter(
    str_detect(generation_technology, "Photovoltaic.*All Technologies") |
    str_detect(generation_technology, "Hydropower.*All Technologies") |
    str_detect(generation_technology, "Wind.*All Technologies") |
    str_detect(generation_technology, "Nuclear.*LWR|Nuclear.*Light Water Reactor") |
    str_detect(generation_technology, "Natural Gas") |
    str_detect(generation_technology, "NGCC") |
    str_detect(generation_technology, "^Oil$|^Oil ") |
    str_detect(generation_technology, "Coal.*All Technologies")
  ) %>%
  mutate(
    tech_name = case_when(
      str_detect(generation_technology, "Photovoltaic.*All") ~ "Photovoltaic",
      str_detect(generation_technology, "Hydropower.*All") ~ "Hydropower",
      str_detect(generation_technology, "Wind.*All") ~ "Wind",
      str_detect(generation_technology, "Nuclear.*LWR|Nuclear.*Light Water") ~ "Nuclear",
      str_detect(generation_technology, "Natural Gas") | str_detect(generation_technology, "NGCC") ~ "Natural Gas",
      str_detect(generation_technology, "^Oil$|^Oil ") ~ "Oil",
      str_detect(generation_technology, "Coal.*All") ~ "Coal",
      TRUE ~ generation_technology
    ),
    tech_type = case_when(
      tech_name == "Hydropower" ~ "Hydropower",
      tech_name %in% c("Coal", "Natural Gas", "Oil") ~ "Fossil",
      TRUE ~ "Other"
    )
  ) %>%
  filter(!is.na(tech_name))
```

# Visualization 1: Slope Graph — Upfront vs. Lifetime (Hydropower Highlighted)

Key technologies from wrangled set. Lines show trajectory from upfront (construction) to lifetime total emissions.

```{r viz8-slope-graph}
key_techs <- gleam_clean %>%
  filter(tech_name %in% c("Hydropower", "Natural Gas", "Oil", "Coal", "Nuclear"))

viz8_data <- key_techs %>%
  filter(!is.na(upstream_med), !is.na(total_med)) %>%
  select(tech_name, tech_type, upfront = upstream_med, lifetime = total_med) %>%
  pivot_longer(cols = c(upfront, lifetime), names_to = "phase", values_to = "emissions") %>%
  mutate(phase = factor(phase, levels = c("upfront", "lifetime")))

viz8 <- viz8_data %>%
  ggplot(aes(x = phase, y = emissions, group = tech_name, color = tech_type)) +
  geom_line(size = 2, alpha = 0.7) +
  geom_point(size = 5, alpha = 0.9, stroke = 2, shape = 21, aes(fill = tech_type)) +
  geom_text_repel(data = filter(viz8_data, phase == "lifetime"),
                  aes(label = tech_name), hjust = -0.1, size = 3.5, fontface = "bold", point.padding = 0.5) +
  annotate("text", x = 1, y = 50, label = "HIGH\nUPFRONT", color = "#2c7bb6", fontface = "bold", size = 4) +
  annotate("text", x = 2, y = 800, label = "FOSSIL FUELS\nWORSE OVER\nLIFETIME", color = "#d73027", fontface = "bold", size = 4) +
  scale_color_manual(values = c("Hydropower" = "#2c7bb6", "Fossil" = "#d73027", "Other" = "#abd9e9")) +
  scale_fill_manual(values = c("Hydropower" = "#2c7bb6", "Fossil" = "#d73027", "Other" = "#abd9e9")) +
  scale_y_log10(labels = scales::number_format()) +
  scale_x_discrete(labels = c("upfront" = "Upfront\n(Construction)", "lifetime" = "Lifetime\nTotal")) +
  labs(title = "Slope Graph: Upfront vs. Lifetime Emissions",
       subtitle = "Hydropower in blue, fossil fuels in red. Steep slopes = they accumulate over time",
       x = "Life Cycle Phase", y = "Emissions (g CO₂e/kWh, log scale)",
       color = "Technology Type", fill = "Technology Type") +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(face = "bold", size = 11))

print(viz8)
```

# Visualization 2: Parallel Coordinates — Multi-Dimensional Profile (Hydropower Highlighted)

Same key technologies across Upfront, Ongoing, and Lifetime dimensions.

```{r viz13-parallel-coordinates}
viz13_data <- gleam_clean %>%
  filter(tech_name %in% c("Hydropower", "Natural Gas", "Oil", "Coal", "Nuclear"),
         !is.na(upstream_med), !is.na(non_combustion_med), !is.na(total_med)) %>%
  select(tech_name, tech_type, Upfront = upstream_med, Ongoing = non_combustion_med, Lifetime = total_med) %>%
  pivot_longer(cols = c(Upfront, Ongoing, Lifetime), names_to = "dimension", values_to = "value") %>%
  mutate(dimension = factor(dimension, levels = c("Upfront", "Ongoing", "Lifetime")))

viz13 <- viz13_data %>%
  ggplot(aes(x = dimension, y = value, group = tech_name, color = tech_type)) +
  geom_line(size = 2, alpha = 0.7) +
  geom_point(size = 4, alpha = 0.9, stroke = 2, shape = 21, aes(fill = tech_type)) +
  geom_text_repel(data = filter(viz13_data, dimension == "Lifetime"),
                  aes(label = tech_name), hjust = -0.1, size = 3, fontface = "bold", point.padding = 0.5) +
  scale_color_manual(values = c("Hydropower" = "#2c7bb6", "Fossil" = "#d73027", "Other" = "#abd9e9")) +
  scale_fill_manual(values = c("Hydropower" = "#2c7bb6", "Fossil" = "#d73027", "Other" = "#abd9e9")) +
  scale_y_log10(labels = scales::number_format()) +
  labs(title = "Parallel Coordinates: Multi-Dimensional Emission Profile",
       subtitle = "Hydropower highlighted. Fossil fuel lines rise across dimensions.",
       x = "Emission Dimension", y = "Emissions (g CO₂e/kWh, log scale)",
       color = "Technology Type", fill = "Technology Type") +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(face = "bold", size = 11))

print(viz13)
```

# Variability 1: Total Life Cycle — Horizontal Boxplot

Boxplot from reported 5-number summary (min, 1Q, median, 3Q, max). Hydropower in blue, fossil fuels in red; other technologies in gray.

```{r variability-boxplot}
box_data <- gleam_clean %>%
  filter(!is.na(total_min), !is.na(total_1q), !is.na(total_med), !is.na(total_3q), !is.na(total_max)) %>%
  group_by(tech_name, tech_type) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(tech_name = fct_reorder(tech_name, total_med, .desc = TRUE))

ggplot(box_data, aes(
  x = tech_name,
  ymin = total_min,
  lower = total_1q,
  middle = total_med,
  upper = total_3q,
  ymax = total_max,
  fill = tech_type,
  color = tech_type
)) +
  geom_boxplot(stat = "identity", width = 0.65, alpha = 0.85, size = 0.9, outlier.shape = NA) +
  coord_flip() +
  scale_fill_manual(
    values = c(Hydropower = "#2c7bb6", Fossil = "#d73027", Other = "gray72"),
    breaks = c("Hydropower", "Fossil", "Other"),
    labels = c("Hydropower", "Fossil Fuels", "Wind/Photovoltaic/Nuclear"),
    name = NULL
  ) +
  scale_color_manual(
    values = c("Hydropower" = "#1a5276", "Fossil" = "#922b21", "Other" = "gray45"),
    guide = "none"
  ) +
  scale_y_log10(labels = scales::number_format()) +
  labs(
    title = "Life Cycle GHG Emissions by Technology",
    subtitle = "Horizontal boxplot from GLEAM 5-number summary (min, 1Q, median, 3Q, max). Hydropower in blue, fossil fuels in red.",
    x = NULL,
    y = "Life Cycle Emissions (g CO₂e/kWh)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40", size = 10)
  ) +
  guides(
    fill = guide_legend(override.aes = list(
      alpha = 1,
      colour = NA
    ))
  )
```

# Variability 1 — Experiment: Layered (overlaid) box-and-whiskers

This version puts **all seven box-and-whiskers on the same left-to-right axis**, layered on top of each other (transparent fill). Use it to see overlap patterns.

```{r variability-boxplot-layered, fig.height=3}
box_exp <- gleam_clean %>%
  filter(!is.na(total_min), !is.na(total_1q), !is.na(total_med), !is.na(total_3q), !is.na(total_max)) %>%
  group_by(tech_name, tech_type) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(draw_order = factor(tech_type, levels = c("Fossil", "Hydropower", "Other"))) %>%
  arrange(draw_order)

type_fill <- c("Hydropower" = "#1a5276", "Fossil" = "#922b21", "Other" = "gray40")
type_col <- c("Hydropower" = "#1a5276", "Fossil" = "#922b21", "Other" = "gray40")

# Overlaid: set a constant category, group by technology, then flip so emissions are left→right
ggplot(box_exp, aes(
  x = "All technologies (layered)",
  group = tech_name,
  ymin = total_min,
  lower = total_1q,
  middle = total_med,
  upper = total_3q,
  ymax = total_max,
  fill = tech_type,
  color = tech_type
)) +
  geom_boxplot(stat = "identity", width = 0.98, alpha = 0.26, size = 1.25, outlier.shape = NA) +
  coord_flip() +
  scale_fill_manual(
    values = type_fill,
    breaks = c("Hydropower", "Fossil", "Other"),
    labels = c("Hydropower", "Fossil Fuels", "Wind/Photovoltaic/Nuclear"),
    name = NULL
  ) +
  scale_color_manual(
    values = type_col,
    guide = "none"
  ) +
  scale_y_log10(labels = scales::number_format()) +
  labs(
    title = "Spread of life-cycle emissions across technologie",
    subtitle = "Renewables have a greater spread and fossil fuels are consistenly high",
    x = NULL,
    y = "Life Cycle Emissions (g CO₂e/kWh)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(3, 8, 3, 8),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(color = "gray40", size = 10, hjust = 0.5)
  ) +
  guides(
    fill = guide_legend(override.aes = list(
      alpha = 1,
      colour = NA
    ))
  )
```

# Summary

The slope graph and parallel coordinates compare key technologies (Coal, Natural Gas, Oil, Nuclear, Hydropower). The variability boxplot shows all seven technologies (Photovoltaic, Hydropower, Wind, Nuclear, Natural Gas, Oil, Coal) using the reported 5-number summary (min, 1Q, median, 3Q, max). The layered boxplot experiment shows how much those distributions overlap when drawn on one shared axis, with hydropower in blue and fossil fuels in red.