---
title: "Hydro Sun"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 6
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
editor_options: 
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 6, fig.height = 6, dpi = 150)
library(tidyverse)
library(haven)
library(lubridate)
```

```{r load}
#| warning: false
#| message: false
dta <- read_dta(here::here("data/marginal-generation-offsets-data.dta"))

inner_radius <- 1.5

inner_month <- dta %>%
  mutate(
    month_num = month(dt),
    month_lab = factor(month.abb[month_num], levels = month.abb)
  ) %>%
  group_by(month_lab) %>%
  summarise(h_mean = mean(h, na.rm = TRUE), .groups = "drop") %>%
  mutate(h_norm = h_mean / max(h_mean, na.rm = TRUE))

month_slots <- tibble(
  month_lab = factor(month.abb, levels = month.abb),
  x_center  = seq(1, 23, by = 2)
)

inner_month_long <- inner_month %>% inner_join(month_slots, by = "month_lab")

outer_hour <- dta %>%
  mutate(hour = hour(dt)) %>%
  group_by(hour) %>%
  summarise(h_mean = mean(h, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    h_norm   = h_mean / max(h_mean, na.rm = TRUE),
    hour_rot = (hour - 12) %% 24,
    inner_r  = inner_radius + 0.1,
    outer_r  = inner_r + 0.9 * h_norm
  )
```

```{r sun-plot}
#| warning: false
#| message: false
label_r <- inner_radius + 1.15
time_labels <- tibble(
  hour_rot = c(0, 6, 12, 18),
  label = c("12pm", "6pm", "12am", "6am"),
  hjust = c(0.5, -0.1, 0.5, 1.1),
  vjust = c(-0.5, 0.5, 1.2, 0.5)
)

month_labels <- tibble(
  x_center = c(1, 7, 13, 19),
  y_pos = inner_radius * 0.6,
  label = c("Jan", "Apr", "Jul", "Oct")
)

ggplot() +
  geom_col(
    data = inner_month_long,
    aes(x = x_center, y = inner_radius, fill = h_norm),
    width = 2, color = NA
  ) +
  geom_segment(
    data = outer_hour,
    aes(x = hour_rot, xend = hour_rot, y = inner_r, yend = outer_r, color = h_norm),
    lineend = "round", linewidth = 1.0, alpha = 0.9
  ) +
  geom_text(data = time_labels, aes(x = hour_rot, y = label_r, label = label),
            hjust = time_labels$hjust, vjust = time_labels$vjust, size = 2.8, color = "gray40") +
  geom_text(data = month_labels, aes(x = x_center, y = y_pos, label = label),
            size = 2.5, color = "gray50") +
  coord_polar(theta = "x", start = 0) +
  scale_colour_gradientn(
    colours = c("#fff7bc", "#fee391", "#fec44f", "#fe9929", "#d95f0e", "#993404"),
    name = "Avg hydropower\n(normalized)"
  ) +
  scale_fill_gradientn(
    colours = c("#fff7bc", "#fee391", "#fec44f", "#fe9929", "#d95f0e", "#993404"),
    guide = "none"
  ) +
  scale_y_continuous(limits = c(0, inner_radius + 1.3), breaks = NULL) +
  labs(
    title = "Hydropower \"sun\": months at the core, hours as corona",
    subtitle = "12:00 at top | Brighter = more generation",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40")
  )
```

### Minimal version (no legend, title, or subtitle)

```{r sun-plot-minimal}
#| warning: false
#| message: false
sun <- ggplot() +
  geom_col(
    data = inner_month_long,
    aes(x = x_center, y = inner_radius, fill = h_norm),
    width = 2, color = NA
  ) +
  geom_segment(
    data = outer_hour,
    aes(x = hour_rot, xend = hour_rot, y = inner_r, yend = outer_r, color = h_norm),
    lineend = "round", linewidth = 1.0, alpha = 0.9
  ) +
  geom_text(data = time_labels, aes(x = hour_rot, y = label_r, label = label),
            hjust = time_labels$hjust, vjust = time_labels$vjust, size = 3.5, color = "#222222") +
  geom_text(data = month_labels, aes(x = x_center, y = y_pos, label = label),
            size = 3.5, color = "#222222") +
  coord_polar(theta = "x", start = 0) +
  scale_colour_gradientn(
    colours = c("#fff7bc", "#fee391", "#fec44f", "#fe9929", "#d95f0e", "#993404"),
    name = NULL,
    guide = guide_colorbar(title = NULL)
  ) +
  scale_fill_gradientn(
    colours = c("#fff7bc", "#fee391", "#fec44f", "#fe9929", "#d95f0e", "#993404"),
    guide = "none"
  ) +
  scale_y_continuous(limits = c(0, inner_radius + 1.3), breaks = NULL) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )

ggsave(
  filename = "sun.pdf",
  plot = sun,
  bg = "transparent")

sun
```

**Takeaway:** Hydropower generation follows clear daily and seasonal patterns — stronger during peak daytime hours and in certain months, which helps grid operators balance supply and demand.

**Data:** Hourly generation from marginal-generation-offsets-data.dta (Platte River Power Authority / Colorado River Storage Project). Averages across months and hours are normalized so brighter = more hydropower output. Regional example, not global.

**Key terms:** Normalized = scaled so the max = 1 (brighter = more output). Marginal generation = power produced when demand is high (e.g., midday). Corona = outer rays; disc = inner circle.
