---
title: "Capacity Doubling"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 6
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 6, fig.height = 5, dpi = 150)
library(tidyverse)
```

```{r load}
#| warning: false
#| message: false
nze <- read_csv(here::here("data/NZE2021_AnnexA.csv"), show_col_types = FALSE)

capacity <- nze %>%
  filter(
    Product == "Hydro",
    Flow == "Installed total power capacity",
    Unit == "GW",
    Year %in% c(2019, 2050)
  ) %>%
  select(Year, Hydropower = Value) %>%
  mutate(
    x = c(0.25, 0.75),
    y = c(0.6, 0.6),
    size = sqrt(Hydropower) / 1.2  # large bubbles (diameter in mm)
  )
```

```{r plot}
#| warning: false
#| message: false
# Two big bubbles: 2019 vs 2050 capacity, one arrow for change
ggplot(capacity, aes(x = x, y = y, size = size)) +
  geom_point(shape = 21, fill = "#1c5d99", color = "#222222", stroke = 1.5) +
  annotate("segment", x = 0.25, xend = 0.75, y = 0.25, yend = 0.25,
           color = "#222222", linewidth = 1.5,
           arrow = arrow(length = unit(0.2, "inches"), type = "closed")) +
  scale_size_identity() +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )
```

**Takeaway:** The IEA Net Zero pathway requires roughly doubling global hydropower capacity by 2050 — from ~1300 GW today to ~2600 GW. Installed capacity is the total power that hydropower plants can produce at full output.

**Data:** IEA Net Zero by 2050 scenario (NZE2021_AnnexA). Installed capacity in GW for Hydro in 2019 and 2050. Bubble size ∝ √capacity.

**Key terms:** GW = gigawatts (1 GW ≈ 1 million kW). Installed capacity = maximum power a plant can produce if running at full output.