---
title: "Renewables Mix"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 6
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
editor_options: 
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 6, fig.height = 5, dpi = 150)
library(tidyverse)
library(scales)
```

```{r load}
#| warning: false
#| message: false
nze <- read_csv(here::here("data/NZE2021_AnnexA.csv"), show_col_types = FALSE)

generation <- nze %>%
  filter(
    Flow == "Electricity generation",
    Unit == "TWh",
    Product %in% c("Hydro", "Solar PV", "Wind", "Renewables")
  ) %>%
  select(Product, Year, Value) %>%
  pivot_wider(names_from = Product, values_from = Value) %>%
  mutate(
    Other_Renewables = Renewables - Hydro - `Solar PV` - Wind,
    Hydro_Share_Renewables = (Hydro / Renewables) * 100
  )

viz_data <- generation %>%
  select(Year, Hydropower = Hydro, `Solar PV`, Wind, Other_Renewables) %>%
  pivot_longer(cols = -Year, names_to = "Source", values_to = "Generation") %>%
  mutate(
    Source = factor(Source, levels = c("Other_Renewables", "Wind", "Solar PV", "Hydropower")),
    Source_Label = case_when(
      Source == "Other_Renewables" ~ "Other",
      Source == "Solar PV" ~ "Solar PV",
      Source == "Wind" ~ "Wind",
      TRUE ~ "Hydropower"
    )
  )
```

```{r plot}
#| warning: false
#| message: false
ggplot(viz_data, aes(x = Year, y = Generation, fill = Source_Label)) +
  geom_area(alpha = 0.85, color = "#ffffff", linewidth = 0.3) +
  geom_line(data = generation, aes(x = Year, y = Hydro), inherit.aes = FALSE,
            color = "#1c5d99", linewidth = 1.5, linetype = "dashed") +
  geom_point(data = generation, aes(x = Year, y = Hydro), inherit.aes = FALSE,
             color = "#1c5d99", size = 3, shape = 21, fill = "#ffffff", stroke = 1.5) +
  annotate("text", x = 2052, y = generation %>% filter(Year == 2050) %>% pull(Hydro),
           label = "Hydro (dotted)", hjust = 0, vjust = 0.5, size = 3, color = "#1c5d99", fontface = "bold") +
  geom_text(data = generation %>% filter(Year == 2050),
            aes(x = 2050, y = Hydro, label = paste0(round(Hydro_Share_Renewables, 0), "% of renewables")),
            inherit.aes = FALSE, hjust = -0.1, vjust = -0.5, size = 3.2, fontface = "bold", color = "#1c5d99") +
  scale_fill_manual(values = c("Hydropower" = "#1c5d99", "Solar PV" = "#639fab", "Wind" = "#bbcde5", "Other" = "#222222"),
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(breaks = c(2019, 2030, 2040, 2050), expand = expansion(mult = c(0, 0.08))) +
  scale_y_continuous(labels = comma_format()) +
  labs(
    title = "Hydro doubles, share declines",
    subtitle = "Stacked area = total renewables. Dotted line = hydropower alone (absolute generation).",
    x = "Year", y = "Electricity (TWh)", fill = NULL,
    caption = "Source: IEA Net Zero by 2050"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 9, color = "gray50")
  )
```

```{r plot-minimal}
#| warning: false
#| message: false
# Minimal: transparent background, gridlines, no dotted line, no axis labels/scales
renew_mix <- ggplot(viz_data, aes(x = Year, y = Generation, fill = Source_Label)) +
  geom_area(alpha = 1, color = "#ffffff", linewidth = 0.3) +
  scale_fill_manual(values = c("Hydropower" = "#1c5d99", "Solar PV" = "#639fab", "Wind" = "#bbcde5", "Other" = "#222222"),
                    guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.08))) +
  scale_y_continuous(labels = NULL, breaks = NULL) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.caption = element_blank(),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    axis.title = element_blank(),
    axis.text = element_blank()
  )

ggsave(
  filename = "renewable_mix_transparent.pdf",
  plot = renew_mix,
  bg = "transparent",

  # Crucial argument for transparent output file
  #width = 6, height = 4, units = "px"
)

```

```{r plot-minimal-transparent}
#| warning: false
#| message: false
# Same as minimal but: remove dotted line, make areas transparent
ggplot(viz_data, aes(x = Year, y = Generation, fill = Source_Label)) +
  geom_area(alpha = 1, color = "#ffffff", linewidth = 0.3) +
  scale_fill_manual(values = c("Hydropower" = "#1c5d99", "Solar PV" = "#639fab", "Wind" = "#bbcde5", "Other" = "#222222"),
                    guide = "none") +
  scale_x_continuous(breaks = c(2019, 2030, 2040, 2050), expand = expansion(mult = c(0, 0.08))) +
  scale_y_continuous(labels = comma_format()) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
  panel.grid.minor = element_blank(),
  plot.title = element_blank(),
  plot.subtitle = element_blank(),
  plot.caption = element_blank(),
  plot.background = element_rect(fill = "transparent", color = NA),
  panel.background = element_rect(fill = "transparent", color = NA)
)
```

**Takeaway:** Hydropower output roughly doubles by 2050, but its share of total renewables falls because solar and wind grow much faster. The dotted line shows hydropower’s absolute path; the hydro band in the stack is the same quantity.

**Data:** IEA Net Zero scenario (NZE2021_AnnexA): electricity generation (TWh) for Hydro, Solar PV, Wind, and total Renewables by year. Other renewables = total renewables minus hydro, solar, and wind.

**Key terms:** TWh = terawatt-hours (1 TWh = 1 billion kWh). Dotted line = absolute hydropower generation; stacked areas = share of total renewables. Renewables = sources that don’t burn fossil fuels (hydro, solar, wind, geothermal, etc.).

