---
title: "Pathway Shift"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 6
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 6, fig.height = 5, dpi = 150)
library(tidyverse)
library(readxl)
library(here)
```

```{r load}
#| warning: false
#| message: false
raw_ts <- read_xlsx(here("data/moesm.xlsx"), sheet = 1)

pathway_cols <- c("CO2_diff", "CH4_diff", "CH4_bub", "CH4_deg")
pathway_labels <- c("CO2 diffusion", "CH4 diffusion", "CH4 ebullition", "CH4 degassing")

emissions_pathway <- raw_ts %>%
  select(year = Year, all_of(pathway_cols)) %>%
  pivot_longer(-year, names_to = "pathway", values_to = "tmol_c") %>%
  mutate(
    pathway = factor(pathway, levels = pathway_cols, labels = pathway_labels),
    pathway = factor(pathway, levels = c("CH4 degassing", "CH4 ebullition", "CH4 diffusion", "CO2 diffusion"))
  )

pathway_fill <- c("CO2 diffusion" = "#1c5d99", "CH4 diffusion" = "#639fab",
                  "CH4 ebullition" = "#bbcde5", "CH4 degassing" = "#222222")
```

```{r plot}
#| warning: false
#| message: false
ggplot(emissions_pathway, aes(x = year, y = tmol_c, fill = pathway)) +
  geom_area(alpha = 0.85, color = "#ffffff", linewidth = 0.2) +
  scale_fill_manual(values = pathway_fill) +
  labs(
    title = "Reservoir emissions shift: CO2 → CH4 dominance",
    subtitle = "CO2 diffusion dominated early; CH4 ebullition & degassing rise as reservoirs age",
    x = "Year", y = "Carbon emissions (Tmol C/yr)", fill = "Pathway",
    caption = "Source: Soued et al. (2022) Nature Geoscience"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 9, color = "gray50")
  )
```

**Takeaway:** Reservoir emissions have shifted from CO2-dominated (early 1900s) to CH4-dominated (today). Methane ebullition (bubbles from sediment) and degassing (from turbines/spillways) now drive most warming impact.

**Data:** Soued et al. (2022) Nature Geoscience, moesm.xlsx. Global reservoir carbon emissions by pathway (CO2 diffusion, CH4 diffusion, CH4 ebullition, CH4 degassing). Tmol C/yr = teramoles of carbon per year.

**Key terms:** Diffusion = gas leaving water surface; ebullition = methane bubbles from sediment; degassing = methane released when water passes through turbines or spillways. CH4 = methane; CO2 = carbon dioxide.