---
title: "Decoupling"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 6
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 6, fig.height = 5, dpi = 150)
library(tidyverse)
library(readxl)
library(here)
```

```{r load}
#| warning: false
#| message: false
raw_ts <- read_xlsx(here("data/moesm.xlsx"), sheet = 1)

totals <- raw_ts %>%
  transmute(
    year = Year,
    tmol_c = CO2_diff + CH4_diff + CH4_bub + CH4_deg,
    rf_mw = RF_tot_rcp6
  ) %>%
  mutate(
    tmol_norm = tmol_c / max(tmol_c, na.rm = TRUE),
    rf_norm = rf_mw / max(rf_mw, na.rm = TRUE)
  ) %>%
  pivot_longer(c(tmol_norm, rf_norm), names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric,
    "tmol_norm" = "Carbon emissions (peaked ~1987)",
    "rf_norm" = "Radiative forcing (keeps rising)"
  ))
```

```{r plot}
#| warning: false
#| message: false
ggplot(totals, aes(x = year, y = value, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_y_continuous(
    name = "Indexed to max (0–1)",
    limits = c(0, 1.05),
    labels = scales::percent_format(scale = 100)
  ) +
  scale_color_manual(
    values = c("Carbon emissions (peaked ~1987)" = "#1c5d99", "Radiative forcing (keeps rising)" = "#639fab"),
    guide = guide_legend(title = NULL)
  ) +
  labs(
    title = "Emissions peaked, radiative forcing did not",
    subtitle = "Both indexed to their max. Emissions fall; RF rises because CH4 lingers in the atmosphere.",
    x = "Year",
    caption = "Source: Soued et al. (2022) Nature Geoscience"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 9, color = "gray50")
  )
```

**Takeaway:** Reservoir carbon emissions peaked around 1987 and have declined, but radiative forcing (the actual warming effect) keeps rising. Methane has a shorter lifetime than CO2 but a much stronger warming effect per molecule, so past methane keeps adding to warming.

**Data:** Soued et al. (2022) Nature Geoscience, moesm.xlsx. Total reservoir carbon emissions (Tmol/yr) and radiative forcing (mW/m²). Both indexed to their max (0–1) so they overlay on the same axis.

**Key terms:** Radiative forcing (RF) = how much extra heat the atmosphere traps due to a gas (measured in mW/m²). Emissions peaked; RF did not because methane lingers in the atmosphere and adds to warming over time.
