---
title: "Good vs Bad CI"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 6
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 6, fig.height = 5, dpi = 150)
library(tidyverse)
```

```{r load}
#| warning: false
#| message: false
library(patchwork)

# Li & He (2022) Table S3
depth_ci <- tribble(
  ~class,                  ~ci,
  "Shallow (<20 m)",       124.5,
  "Medium (20–100 m)",      66.0,
  "Deep (>100 m)",          27.7
)

trophic_ci <- tribble(
  ~class,            ~ci,
  "Oligotrophic",     55.6,
  "Mesotrophic",      69.1,
  "Eutrophic",       116.7
)
```

```{r plot}
#| warning: false
#| message: false
p1 <- ggplot(depth_ci, aes(x = fct_reorder(class, ci), y = ci)) +
  geom_col(fill = "#1c5d99", width = 0.6, alpha = 0.85) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "#222222", linewidth = 0.8) +
  coord_flip() +
  labs(subtitle = "Water depth", x = NULL, y = "kg CO₂e/MWh") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_blank())

p2 <- ggplot(trophic_ci, aes(x = fct_reorder(class, ci), y = ci)) +
  geom_col(fill = "#639fab", width = 0.6, alpha = 0.85) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "#222222", linewidth = 0.8) +
  coord_flip() +
  labs(subtitle = "Trophic level (nutrients)", x = NULL, y = "kg CO₂e/MWh") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_blank())

(p1 | p2) +
  plot_annotation(
    title = "Planning matters: depth & trophic",
    subtitle = "Dotted red line = 80 kg CO₂e/MWh (low-carbon target). Bars above = fail target.",
    caption = "Source: Li & He (2022) | CI = carbon intensity (emissions per MWh generated)"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    plot.caption = element_text(size = 8, color = "gray50")
  )
```

```{r plot_grid}
#| warning: false
#| message: false
# Simple CI “risk strips” for depth and trophic level

depth_heat <- depth_ci %>%
  mutate(
    depth_simple = case_when(
      str_detect(class, "Shallow") ~ "Shallow",
      str_detect(class, "Medium")  ~ "Medium",
      str_detect(class, "Deep")    ~ "Deep",
      TRUE ~ class
    )
  ) %>%
  ggplot(aes(x = depth_simple, y = 1, fill = ci)) +
  geom_tile(color = "white", linewidth = 0.7) +
  geom_text(aes(label = round(ci)), color = "white", fontface = "bold") +
  scale_fill_gradient2(
    low = "#1c5d99", mid = "#bbcde5", high = "#b8552e",
    midpoint = 80,
    name = "Median CI\n(kg CO₂e/MWh)"
  ) +
  labs(
    title = "By depth",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

trophic_heat <- trophic_ci %>%
  ggplot(aes(x = class, y = 1, fill = ci)) +
  geom_tile(color = "white", linewidth = 0.7) +
  geom_text(aes(label = round(ci)), color = "white", fontface = "bold") +
  scale_fill_gradient2(
    low = "#1c5d99", mid = "#bbcde5", high = "#b8552e",
    midpoint = 80,
    name = "Median CI\n(kg CO₂e/MWh)"
  ) +
  labs(
    title = "By nutrient level (trophic)",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )

depth_heat / trophic_heat +
  plot_annotation(
    title = "Which reservoirs are high‑carbon?",
    subtitle = "Li & He (2022): shallow and eutrophic classes cluster above the 80 kg CO₂e/MWh low‑carbon target."
  )
```

**Takeaway:** Shallow reservoirs (&lt;20 m) and eutrophic (nutrient-rich) reservoirs have much higher carbon intensity than deep or oligotrophic ones. Siting and management matter: planning for depth and water quality can keep hydropower low-carbon.

**Data:** Li & He (2022) — median carbon intensity (CI) by depth class and trophic level from supplementary Table S3. **CI = carbon intensity**: kg of CO₂-equivalent emissions per MWh of electricity generated. The **dotted red line (80 kg CO₂e/MWh)** is the UN 2030 Agenda low-carbon threshold; reservoirs above it are not considered “low-carbon.” Trophic = nutrient level (oligo = low, eutrophic = high).
