---
title: "Share Failing 80 kg"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 6
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 6, fig.height = 5, dpi = 150)
library(tidyverse)
library(here)
```

```{r load}
#| warning: false
#| message: false
# Share failing 80 kg CO2e/MWh — Li & He (2022), from GRanD + FHReD dam databases
ci_existing_future <- read_csv(here("data/share_failing_80kg.csv"), show_col_types = FALSE)

share_failing <- ci_existing_future %>%
  pivot_longer(cols = c(pct_above_80, pct_below_80),
               names_to = "category", values_to = "percent") %>%
  mutate(
    category = recode(category,
      "pct_above_80" = ">\n80 kg CO₂e/MWh",
      "pct_below_80" = "≤\n80 kg CO₂e/MWh"
    )
  )
```

```{r plot}
#| warning: false
#| message: false
ggplot(share_failing, aes(x = group, y = percent, fill = category)) +
  geom_col(width = 0.6, color = "#ffffff", linewidth = 0.5) +
  scale_fill_manual(values = c(">\n80 kg CO₂e/MWh" = "#222222", "≤\n80 kg CO₂e/MWh" = "#bbcde5")) +
  labs(
    title = "Who meets the low-carbon target?",
    subtitle = "Carbon target = 80 kg CO₂e/MWh (UN 2030 Agenda). Light = meets target; dark = exceeds it.",
    x = NULL, y = "% of reservoirs", fill = NULL,
    caption = "Source: Li & He (2022), from GRanD + FHReD dam databases"
  ) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    legend.position = "bottom",
    plot.caption = element_text(size = 9, color = "gray50")
  )
```

```{r plot-plain}
#| warning: false
#| message: false
# Plain version: minimal, no title/subtitle/caption/legend/axes/group labels
ggplot(share_failing, aes(x = group, y = percent, fill = category)) +
  geom_col(width = 0.6, color = "#ffffff", linewidth = 0.5) +
  scale_fill_manual(values = c(">\n80 kg CO₂e/MWh" = "#222222", "≤\n80 kg CO₂e/MWh" = "#bbcde5"), guide = "none") +
  labs(x = NULL, y = NULL) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.caption = element_blank(),
    axis.text.y = element_blank()
  )
```

**Takeaway:** About 56% of existing dams stay below the 80 kg CO₂e/MWh low-carbon threshold, but only 34% of planned dams do. Future expansion is tilting toward higher-carbon reservoirs unless siting and design improve.

**Data:** Li & He (2022), derived from GRanD (existing dams) + FHReD (future dams). Share of reservoirs exceeding 80 kg CO₂e/MWh. **Carbon target (80 kg CO₂e/MWh):** set by the UN 2030 Agenda; power sources below this are considered “low-carbon.” Dams above it emit more CO₂-equivalent per MWh than this benchmark.
