---
title: "Future Hydropower Pipeline — Alluvial"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 8
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 8, fig.height = 6, dpi = 150)
library(tidyverse)
library(here)
library(ggalluvial)
```

```{r load}
#| warning: false
#| message: false
# Future hydropower dams (FHReD, Zarfl et al. 2015)
future <- readxl::read_xlsx(
  here("data/FHReD_2015_future_dams_Zarfl_et_al_beta_version/FHReD_2015_future_dams_Zarfl_et_al_beta_version.xlsx"),
  sheet = 1
)

pal <- c(
  carbon_black = "#222222",
  white        = "#ffffff",
  baltic_blue  = "#1c5d99",
  pacific_blue = "#639fab",
  powder_blue  = "#bbcde5"
)

# Keep top groups by MW (not by row count), then build alluvial table
continent_mw <- future %>%
  filter(!is.na(`Capacity (MW)`), !is.na(Continent)) %>%
  group_by(Continent) %>%
  summarise(mw = sum(`Capacity (MW)`, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mw))

top_continents <- continent_mw %>% slice_head(n = 5) %>% pull(Continent)

basin_mw <- future %>%
  filter(!is.na(`Capacity (MW)`), !is.na(`Major Basin`)) %>%
  group_by(`Major Basin`) %>%
  summarise(mw = sum(`Capacity (MW)`, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mw))

top_basins <- basin_mw %>% slice_head(n = 5) %>% pull(`Major Basin`)

future_alluvial <- future %>%
  filter(!is.na(`Capacity (MW)`)) %>%
  mutate(
    continent  = if_else(Continent %in% top_continents, Continent, "Other"),
    basin_group = if_else(`Major Basin` %in% top_basins, `Major Basin`, "Other basins")
  ) %>%
  # Size classes for installed capacity (3 buckets to reduce clutter)
  mutate(
    size_class = cut(
      `Capacity (MW)`,
      breaks = c(-Inf, 100, 500, Inf),
      labels = c("<100 MW", "100–500 MW", ">500 MW"),
      right = FALSE
    )
  ) %>%
  # Summarise flows: one row per combination of axes
  group_by(continent, basin_group, size_class) %>%
  summarise(
    mw_total = sum(`Capacity (MW)`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(mw_total > 0)
```

```{r plot_vertical_alluvial}
#| warning: false
#| message: false
# Vertical alluvial, weighted by MW

ggplot(
  future_alluvial,
  aes(
    axis1 = size_class,
    axis2 = basin_group,
    axis3 = continent,
    y     = mw_total
  )
) +
  geom_alluvium(
    aes(fill = size_class),
    alpha = 0.9,
    knot.pos = 0.45,
    color = NA
  ) +
  # Invisible strata just to anchor labels (no boxes)
  geom_stratum(fill = NA, color = NA, width = 0.25) +
  geom_text(
    stat = "stratum",
    aes(label = stringr::str_wrap(after_stat(stratum), width = 14)),
    size = 2.8,
    lineheight = 0.9,
    color = pal["carbon_black"],
    check_overlap = TRUE
  ) +
  scale_fill_manual(
    values = c(
      "<100 MW"    = pal["powder_blue"],
      "100–500 MW" = pal["pacific_blue"],
      ">500 MW"    = pal["baltic_blue"]
    )
  ) +
  scale_x_discrete(
    limits = c("Size", "Basin", "Region"),
    expand = c(0.04, 0.04)
  ) +
  labs(
    title    = "Future hydropower pipeline",
    subtitle = "Width = capacity (MW). Flows show how projects distribute by dam size, river basin, and region.",
    x        = NULL,
    y        = "Capacity (MW)",
    fill     = "Dam size"
  ) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title   = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9, color = "gray40"),
    plot.margin = margin(6, 6, 6, 6),
    legend.position = "right"
  )
```

**Takeaway:** The global hydropower pipeline is heavily concentrated in a limited set of continents and river basins, and a large share of new capacity comes from very large and mega-projects. Because these projects will lock in infrastructure and emissions for decades, careful siting and design are critical for keeping future hydropower truly low-carbon.

**Data:** FHReD (Future Hydropower Reservoirs and Dams; Zarfl et al. 2015). Each ribbon is weighted by installed capacity (MW) from FHReD. Categories are derived from project stage, continent, major basin, and capacity class.

