---
title: "Future Hydropower — Where It's Going"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 8
    fig-height: 10
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, dpi = 150)
library(tidyverse)
library(here)
library(readxl)
library(ggalluvial)
library(treemapify)
library(maps)
library(scales)
```

```{r load}
#| warning: false
#| message: false
# Future hydropower dams (FHReD, Zarfl et al. 2015)
future <- read_xlsx(
  here("data/FHReD_2015_future_dams_Zarfl_et_al_beta_version/FHReD_2015_future_dams_Zarfl_et_al_beta_version.xlsx"),
  sheet = 2
)

pal <- c(
  carbon_black = "#222222",
  white        = "#ffffff",
  baltic_blue  = "#1c5d99",
  pacific_blue = "#639fab",
  powder_blue  = "#bbcde5"
)

# Clean and prepare
future_geo <- future %>%
  filter(!is.na(`Capacity (MW)`), !is.na(LAT_cleaned), !is.na(Lon_Cleaned)) %>%
  mutate(
    continent = coalesce(Continent, "Other"),
    basin = coalesce(`Major Basin`, "Other")
  )

# Top basins for Sankey and treemap
basin_mw <- future_geo %>%
  group_by(basin) %>%
  summarise(mw = sum(`Capacity (MW)`), .groups = "drop") %>%
  arrange(desc(mw))

top_basins <- basin_mw %>% slice_head(n = 6) %>% pull(basin)

future_geo <- future_geo %>%
  mutate(
    basin_group = if_else(basin %in% top_basins, basin, "Other basins"),
    basin_group = forcats::fct_relevel(basin_group, "Other basins", after = Inf)
  )
```

## 1. Map — Where future dams are located

```{r map}
#| warning: false
#| message: false
# World map outline (requires maps package)
world <- map_data("world")

p_map <- ggplot() +
  geom_polygon(
    data = world,
    aes(x = long, y = lat, group = group),
    fill = "#f5f5f5", color = "#dddddd", linewidth = 0.2
  ) +
  geom_point(
    data = future_geo,
    aes(x = Lon_Cleaned, y = LAT_cleaned, size = `Capacity (MW)`, color = continent),
    alpha = 0.6,
    shape = 16
  ) +
  scale_size_continuous(range = c(1, 10), labels = comma) +
  scale_color_manual(
    values = c(
      "Africa" = pal["baltic_blue"], "Asia" = pal["pacific_blue"],
      "South America" = pal["powder_blue"], "North America" = "#94b3d7",
      "Europe" = "#c7d7f2", "Oceania" = "#e0ecf4", "Other" = pal["carbon_black"]
    )
  ) +
  coord_fixed(ratio = 1.3, xlim = c(-180, 180), ylim = c(-60, 75)) +
  labs(
    title = "Future hydropower dam locations",
    subtitle = "Point size = capacity (MW). Color = continent.",
    x = NULL, y = NULL, size = "MW", color = "Continent"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

p_map
```

## 2. Trunk + branches — Capacity flows by basin

```{r sankey}
#| warning: false
#| message: false
basin_tab <- future_geo %>%
  group_by(basin_group) %>%
  summarise(mw = sum(`Capacity (MW)`), .groups = "drop") %>%
  arrange(desc(mw)) %>%
  mutate(
    basin_group = forcats::fct_relevel(basin_group, "Other basins", after = Inf)
  ) %>%
  mutate(
    total_node = "Total future hydropower",
    share = mw / sum(mw)
  )

ggplot(basin_tab, aes(axis1 = total_node, axis2 = basin_group, y = mw)) +
  geom_alluvium(aes(fill = basin_group), width = 0, alpha = 0.9, knot.pos = 0.4) +
  geom_stratum(width = 0.2, fill = pal["white"], color = "#e5e5e5") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3, color = pal["carbon_black"]) +
  scale_fill_brewer(palette = "Blues", guide = "none") +
  scale_x_discrete(limits = c("Total", "River basin"), expand = c(0.1, 0.1)) +
  labs(title = "Capacity by river basin", subtitle = "Width = MW", x = NULL, y = "MW") +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 3. Bar chart — MW by continent

```{r bar}
#| warning: false
#| message: false
cont_mw <- future_geo %>%
  group_by(continent) %>%
  summarise(mw = sum(`Capacity (MW)`), .groups = "drop") %>%
  arrange(desc(mw))

p_bar <- ggplot(cont_mw, aes(x = fct_reorder(continent, mw), y = mw, fill = continent)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = comma(round(mw))), hjust = -0.1, size = 3, color = pal["carbon_black"]) +
  scale_fill_manual(
    values = c(
      "Africa" = pal["baltic_blue"], "Asia" = pal["pacific_blue"],
      "South America" = pal["powder_blue"], "North America" = "#94b3d7",
      "Europe" = "#c7d7f2", "Oceania" = "#e0ecf4", "Other" = pal["carbon_black"]
    ),
    guide = "none"
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(
    title = "Future capacity by continent",
    x = NULL, y = "Capacity (MW)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold")
  )

p_bar
```

## 4. Treemap — Continent → Basin → MW

```{r treemap}
#| warning: false
#| message: false
treemap_data <- future_geo %>%
  group_by(continent, basin_group) %>%
  summarise(mw = sum(`Capacity (MW)`), .groups = "drop") %>%
  filter(mw > 0)

p_treemap <- ggplot(treemap_data, aes(area = mw, fill = continent, subgroup = continent, label = basin_group)) +
  geom_treemap(layout = "squarified", color = "white", linewidth = 0.5) +
  geom_treemap_subgroup_border(layout = "squarified", color = pal["carbon_black"], linewidth = 1) +
  geom_treemap_text(
    layout = "squarified", place = "centre",
    size = 10, color = "white", fontface = "bold",
    min.size = 4
  ) +
  scale_fill_manual(
    values = c(
      "Africa" = pal["baltic_blue"], "Asia" = pal["pacific_blue"],
      "South America" = pal["powder_blue"], "North America" = "#94b3d7",
      "Europe" = "#c7d7f2", "Oceania" = "#e0ecf4", "Other" = pal["carbon_black"]
    ),
    guide = "legend"
  ) +
  labs(title = "Capacity by continent and basin", fill = "Continent") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

p_treemap
```

---

**Takeaway:** Future hydropower capacity is concentrated in Asia, Africa, and South America, and within a small number of major river basins (Yangtze, Amazon, Congo, Indus, Ganges-Brahmaputra). Planning in these regions and basins will determine whether that expansion is low-impact.

**Data:** FHReD (Future Hydropower Reservoirs and Dams; Zarfl et al. 2015).
