---
title: "Future Hydropower — Where It's Going"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 8
    fig-height: 10
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, dpi = 150)
library(tidyverse)
library(here)
library(readxl)
library(ggalluvial)
library(treemapify)
library(maps)
library(scales)
source(here("final_graphics/style_sheet.R"))
```

```{r load}
#| warning: false
#| message: false
# Future hydropower dams (FHReD, Zarfl et al. 2015)
future <- read_xlsx(
  here("data/FHReD_2015_future_dams_Zarfl_et_al_beta_version/FHReD_2015_future_dams_Zarfl_et_al_beta_version.xlsx"),
  sheet = 2
)

pal <- c(
  carbon_black = "#222222",
  white        = "#ffffff",
  baltic_blue  = "#1c5d99",
  pacific_blue = "#639fab",
  powder_blue  = "#bbcde5"
)

# Clean and prepare
future_geo <- future %>%
  filter(!is.na(`Capacity (MW)`), !is.na(LAT_cleaned), !is.na(Lon_Cleaned)) %>%
  mutate(
    continent = coalesce(Continent, "Other"),
    basin = coalesce(`Major Basin`, "Other")
  )

# Top basins for Sankey and treemap
basin_mw <- future_geo %>%
  group_by(basin) %>%
  summarise(mw = sum(`Capacity (MW)`), .groups = "drop") %>%
  arrange(desc(mw))

top_basins <- basin_mw %>% slice_head(n = 10) %>% pull(basin)

future_geo <- future_geo %>%
  mutate(
    basin_group = if_else(basin %in% top_basins, basin, "Other basins"),
    basin_group = forcats::fct_relevel(basin_group, "Other basins", after = Inf)
  )

# Match 17's basin palette (top 10 + other) — stable mapping by name
pal_basin10 <- make_top10_basin_palette(top_basins)

# Add ranked basin labels for consistent ordering/labels
future_geo <- future_geo %>%
  mutate(
    basin_ranked = if_else(basin %in% top_basins,
                           paste0(match(basin, top_basins), ". ", basin),
                           "Other basins"),
    basin_ranked = factor(basin_ranked, levels = names(pal_basin10))
  )
```

## 1. Trunk + branches — Capacity flows by basin (top 10 + other)

```{r sankey}
#| warning: false
#| message: false
basin_tab <- future_geo %>%
  group_by(basin_ranked) %>%
  summarise(mw = sum(`Capacity (MW)`), .groups = "drop") %>%
  arrange(desc(mw)) %>%
  mutate(
    total_node = "Total future hydropower",
    share = mw / sum(mw)
  )

ggplot(basin_tab, aes(axis1 = total_node, axis2 = basin_ranked, y = mw)) +
  geom_alluvium(aes(fill = basin_ranked), width = 0, alpha = 0.9, knot.pos = 0.4) +
  geom_stratum(width = 0.2, fill = pal["white"], color = "#e5e5e5") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3, color = pal["carbon_black"]) +
  scale_fill_manual(values = pal_basin10, guide = "none") +
  scale_x_discrete(limits = c("Total", "River basin"), expand = c(0.1, 0.1)) +
  labs(title = "Capacity by river basin", subtitle = "Width = MW", x = NULL, y = "MW") +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 2. Treemap — Capacity by basin (top 10 + other)

```{r treemap}
#| warning: false
#| message: false
treemap_data <- future_geo %>%
  group_by(basin_ranked) %>%
  summarise(mw = sum(`Capacity (MW)`), .groups = "drop") %>%
  filter(mw > 0) %>%
  mutate(label = paste0(basin_ranked, "\n(", scales::comma(round(mw)), " MW)"))

p_treemap <- ggplot(treemap_data, aes(area = mw, fill = basin_ranked, label = label)) +
  geom_treemap(layout = "squarified", color = "white", linewidth = 0.5) +
  geom_treemap_text(
    layout = "squarified", place = "centre",
    size = 10, color = "white", fontface = "bold",
    min.size = 4
  ) +
  scale_fill_manual(values = okabe_ito_11, guide = "none") +
  labs(title = "Capacity by basin (top 10 + other)") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

p_treemap
```

## 3. Basin bubbles — Total future vs top 10 basins

```{r basin-bubbles}
#| warning: false
#| message: false
#| fig-height: 3
# Use the same top-10 basin ordering and colors as the sankey/treemap.

basin_top10 <- basin_mw %>%
  slice_head(n = 10) %>%
  mutate(
    basin_rank = row_number(),
    basin_group = paste0(basin_rank, ". ", basin),
    basin_group = factor(basin_group, levels = paste0(1:10, ". ", top_basins)),
    cap_k = round(mw / 1e3),
    cap_label = paste0(cap_k, "k\nMW")
  )

total_future <- tibble(
  basin_group = "Total future",
  mw = sum(basin_mw$mw),
  cap_k = round(mw / 1e3),
  cap_label = paste0(cap_k, "k\nMW")
)

basin_bubbles <- bind_rows(
  total_future,
  basin_top10 %>% transmute(basin_group = as.character(basin_group), mw, cap_label)
) %>%
  mutate(
    basin_group = factor(basin_group, levels = c("Total future", as.character(basin_top10$basin_group))),
    x_pos = seq(1, by = 0.85, length.out = n())
  )

pal_bubbles <- c(
  "Total future" = "#ffffff",
  setNames(pal_basin10[1:10], as.character(basin_top10$basin_group))
)

ggplot(
  basin_bubbles,
  aes(
    x = x_pos,
    y = 0.9,
    size = mw,
    fill = basin_group
  )
) +
  geom_point(shape = 21, color = pal["carbon_black"], alpha = 0.95, stroke = 0.4) +
  # Two-pass text for legibility (fake outline)
  geom_text(aes(label = cap_label), color = pal["carbon_black"], fontface = "bold", size = 2.9, lineheight = 0.9) +
  geom_text(aes(label = cap_label), color = pal["white"], fontface = "bold", size = 2.6, lineheight = 0.9) +
  scale_fill_manual(values = pal_bubbles, guide = "none") +
  scale_size_continuous(range = c(8, 30), labels = comma, guide = "none") +
  scale_x_continuous(
    breaks = basin_bubbles$x_pos,
    labels = as.character(basin_bubbles$basin_group),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  coord_cartesian(ylim = c(0, 1.6), clip = "off") +
  labs(
    title = "Total future capacity vs top 10 basins",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_text(size = 8, face = "bold", margin = margin(t = 2)),
    plot.title = element_text(face = "bold", hjust = 0)
  )
```

---

**Takeaway:** Future hydropower capacity is concentrated in Asia, Africa, and South America, and within a small number of major river basins (Yangtze, Amazon, Congo, Indus, Ganges-Brahmaputra). Planning in these regions and basins will determine whether that expansion is low-impact.

**Data:** FHReD (Future Hydropower Reservoirs and Dams; Zarfl et al. 2015).
