---
title: "Where Are Future Hydropower Dams Planned?"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
library(tidyverse)
library(sf)
library(readxl)
library(here)
library(scales)
library(rnaturalearth)
library(rnaturalearthdata)
```

```{r load-grand}
#| warning: false
#| message: false
# Existing hydropower reservoirs — GRanD v1.3
grand_path <- here("data/GRanD_Version_1_3/GRanD_dams_v1_3.shp")
grand_raw <- st_read(grand_path, quiet = TRUE)

# Filter to dams with hydropower as a use
# (MAIN_USE == "Hydroelectricity" or electricity listed in USE_ELEC)
grand_hydro <- grand_raw %>%
  st_drop_geometry() %>%
  filter(
    MAIN_USE == "Hydroelectricity" |
      str_detect(coalesce(USE_ELEC, ""), "Main|Sec")
  )

existing_country <- grand_hydro %>%
  group_by(country = COUNTRY) %>%
  summarise(
    existing_dams = n(),
    existing_storage_mcm = sum(replace_na(CAP_MCM, 0)),  # reservoir volume, not power
    .groups = "drop"
  )
```

```{r load-fhred}
#| warning: false
#| message: false
# Future hydropower dams — FHReD (same sheet as other FHReD graphics)
fhred <- read_xlsx(
  here("data/FHReD_2015_future_dams_Zarfl_et_al_beta_version/FHReD_2015_future_dams_Zarfl_et_al_beta_version.xlsx"),
  sheet = "Future_dams_beta"
)

future_country <- fhred %>%
  filter(!is.na(`Capacity (MW)`), !is.na(Country)) %>%
  group_by(country = Country) %>%
  summarise(
    future_dams = n(),
    future_mw   = sum(`Capacity (MW)`, na.rm = TRUE),
    .groups     = "drop"
  )
```

```{r combine}
#| warning: false
#| message: false
country_hydro <- full_join(existing_country, future_country, by = "country") %>%
  mutate(
    existing_dams = replace_na(existing_dams, 0L),
    future_dams   = replace_na(future_dams, 0L),
    future_mw     = replace_na(future_mw, 0)
  )

# Harmonize country names for join with rnaturalearth
country_hydro_clean <- country_hydro %>%
  mutate(
    country_join = case_when(
      # FHReD-specific mismatches
      country == "Bosnia and Herzegovina"   ~ "Bosnia and Herz.",
      country == "Central African Republic" ~ "Central African Rep.",
      country == "Congo, Rep."             ~ "Congo",
      country == "Egypt, Arab Rep."        ~ "Egypt",
      country == "Korea, Dem. Rep."        ~ "North Korea",
      country == "Lao PDR"                 ~ "Laos",
      country == "Macedonia"               ~ "North Macedonia",
      country == "United States"           ~ "United States of America",
      # GRanD-specific mismatches
      country == "Congo (DRC)"             ~ "Dem. Rep. Congo",
      country == "Czech Republic"          ~ "Czechia",
      country == "Dominican Republic"      ~ "Dominican Rep.",
      country == "French Guiana"           ~ "France",
      country == "Ivory Coast"             ~ "Côte d'Ivoire",
      country == "Myanmar (Burma)"         ~ "Myanmar",
      country == "Swaziland"               ~ "eSwatini",
      TRUE ~ country
    )
  )
```

```{r map-data}
#| warning: false
#| message: false
world <- ne_countries(scale = "medium", returnclass = "sf")

# Join by country name (after harmonization)
map_data <- world %>%
  left_join(country_hydro_clean, by = c("name" = "country_join"))
```

```{r map}
#| warning: false
#| message: false
ggplot(map_data) +
  geom_sf(aes(fill = future_dams)) +
  scale_fill_viridis_c(
    trans = "sqrt",
    na.value = "grey95",
    name = "Future hydropower\ndams (count)"
  ) +
  labs(
    title = "Where are future hydropower dams planned?",
    subtitle = "Country-level count of planned hydropower dams (FHReD).",
    caption = "Future dams: FHReD (Zarfl et al. 2015). Existing dams (GRanD v1.3) provide context for current fleets."
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

**Takeaway:** Some countries have very large pipelines of planned hydropower dams, while others have few or none. This map highlights where **future hydropower development is most concentrated** in terms of the number of projects.

**Data:** Future hydropower dams from FHReD (Zarfl et al. 2015), aggregated by country. This map uses the **number of dams** (not generation). Existing hydropower fleets from GRanD v1.3 are used only for context elsewhere.

