---
title: "Future Hydropower — Basins, Biomes, and Countries"
format:
  html:
    self-contained: true
  pdf:
    fig-width: 11
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
---

```{r setup}
#| warning: false
#| message: false
knitr::opts_chunk$set(fig.width = 11, fig.height = 6, dpi = 150)
library(tidyverse)
library(here)
library(readxl)
library(scales)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggnewscale)
library(ggrepel)
source(here("final_graphics/style_sheet.R"))
```

```{r load-fhred}
#| warning: false
#| message: false
# FHReD future dams (same source as 14/15; use sheet 2 if sheet 1 is intro)
xl_path <- here("data/FHReD_2015_future_dams_Zarfl_et_al_beta_version/FHReD_2015_future_dams_Zarfl_et_al_beta_version.xlsx")
future <- read_xlsx(xl_path, sheet = 2)

# Li & He (2022) Table S2 — median carbon intensity by biome (kg CO2e/MWh)
biome_ci <- tribble(
  ~biome,     ~ci_median,
  "Tundra",      43.8,
  "Boreal",     104.4,
  "Temperate",   54.8,
  "Tropical",    91.0
)

assign_biome <- function(lat) {
  case_when(
    lat >= 60 | lat <= -60 ~ "Tundra",
    (lat >= 50 & lat < 60) | (lat <= -50 & lat > -60) ~ "Boreal",
    (lat >= 23 & lat < 50) | (lat <= -23 & lat > -50) ~ "Temperate",
    lat > -23 & lat < 23 ~ "Tropical",
    TRUE ~ NA_character_
  )
}

future_geo <- future %>%
  filter(!is.na(`Capacity (MW)`), !is.na(LAT_cleaned), !is.na(Lon_Cleaned)) %>%
  mutate(
    biome = assign_biome(LAT_cleaned),
    basin = coalesce(`Major Basin`, "Other")
  ) %>%
  left_join(biome_ci, by = "biome")

# Basin-level totals (MW) and top basins for discrete coloring (ordered by size)
basin_mw <- future_geo %>%
  group_by(basin) %>%
  summarise(basin_mw_total = sum(`Capacity (MW)`), .groups = "drop") %>%
  arrange(desc(basin_mw_total)) %>%
  mutate(
    rank = row_number(),
    basin_ranked = if_else(rank <= 10,
                           paste0(rank, ". ", basin),
                           "Other basins")
  )

# Stable top-10 basin palette keyed by ranked basin labels (must be defined before use)
top_basins <- basin_mw %>% slice_head(n = 10) %>% pull(basin)
pal_basin10 <- make_top10_basin_palette(top_basins)

# Label anchors for top 10 basins: mean dam longitude/latitude per basin.
# We then nudge some labels away from crowded regions (e.g., Southeast Asia)
# and draw curved leader lines back to the anchor point.
basin_labels <- future_geo %>%
  filter(basin %in% basin_mw$basin[1:10]) %>%
  group_by(basin) %>%
  summarise(
    # Use medians (not means) so a few outlier dams don't pull the anchor away
    anchor_x = median(Lon_Cleaned, na.rm = TRUE),
    anchor_y = median(LAT_cleaned, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    basin_mw %>%
      filter(rank <= 10) %>%
      transmute(
        basin,
        basin_mw_total,
        label_rank = paste0(rank, ". ", basin)
      ),
    by = "basin"
  ) %>%
  mutate(
    # Simple label: rank + basin only
    label = label_rank,
    # Keep anchors on-basin (for stems), but nudge label placement toward ocean.
    nudge_x = case_when( # + for east, - for west
      basin == "Yangtze"  ~ 60,   
      basin == "Mekong"   ~ 40,  
      basin == "Amazon"   ~ 30, 
      basin == "La Plata" ~ 25,   
      basin == "Congo"    ~ -30, 
      basin == "Nile"     ~ 20,  
      basin == "Irrawaddy" ~ -15,   
      basin == "Salween"  ~ 40, 
      basin == "Indus"    ~ -10,
      basin == "Ganges - Brahmaputra"   ~ 10,
      TRUE ~ 0
    ),
    nudge_y = case_when( # + for north, - for south
      basin == "Yangtze"  ~ 10,  
      basin == "Mekong"   ~ -15,   
      basin == "Amazon"   ~ 10,    
      basin == "La Plata" ~ -5,    
      basin == "Congo"    ~ -15,    
      basin == "Nile"     ~ -5,    
      basin == "Irrawaddy" ~ -25, 
      basin == "Salween"  ~ -5, 
      basin == "Indus"    ~ 10,
      basin == "Ganges - Brahmaputra"   ~ 25,
      TRUE ~ 0
    )
  )

future_geo <- future_geo %>%
  left_join(basin_mw, by = "basin") %>%
  mutate(
    basin_group = factor(
      basin_ranked,
      levels = names(pal_basin10)
    )
  )

# Lat-band rectangles for background (biome zones)
lat_bands <- tribble(
  ~ymin, ~ymax, ~biome,
  -60,   -50,   "Boreal",
  -50,   -23,   "Temperate",
  -23,    23,   "Tropical",
   23,    50,   "Temperate",
   50,    60,   "Boreal"
) %>%
  left_join(biome_ci, by = "biome")

pal_ci <- c(
  # Biome band colors (green ramp): Boreal ~ white, Temperate light green, Tropical darker (but still light) green
  "Boreal"    = "#f5fbf6",
  "Temperate" = "#cfead2",
  "Tropical"  = "#76c58a"
)

# Basin capacity palette (Greys) so basin colors are neutral
pal_capacity <- RColorBrewer::brewer.pal(9, "Greys")
```

```{r load-country-future}
#| warning: false
#| message: false
# Basin polygons (BasinATLAS level 2) + aggregate FHReD dams into those basins
basins_lev02 <- readRDS(here("data/basins_lev02_geom.rds"))
basins_lev02 <- sf::st_as_sf(basins_lev02)
if (is.na(sf::st_crs(basins_lev02))) sf::st_crs(basins_lev02) <- 4326

# BasinATLAS polygons can include invalid rings; fix before spatial join.
sf::sf_use_s2(FALSE)
basins_lev02 <- sf::st_make_valid(basins_lev02)
basins_lev02 <- sf::st_collection_extract(basins_lev02, "POLYGON")
basins_lev02 <- sf::st_cast(basins_lev02, "MULTIPOLYGON", warn = FALSE)

future_pts <- future_geo %>%
  st_as_sf(coords = c("Lon_Cleaned", "LAT_cleaned"), crs = 4326, remove = FALSE)

future_with_basin <- st_join(
  future_pts,
  basins_lev02 %>% select(hybas_id),
  left = FALSE
)

# Restore s2 for downstream ops
sf::sf_use_s2(TRUE)

basin_stats <- future_with_basin %>%
  st_drop_geometry() %>%
  group_by(hybas_id) %>%
  summarise(
    future_dams = n(),
    future_mw   = sum(`Capacity (MW)`, na.rm = TRUE),
    .groups     = "drop"
  )

basin_map <- basins_lev02 %>%
  left_join(basin_stats, by = "hybas_id") %>%
  mutate(
    future_dams = replace_na(future_dams, 0L),
    future_mw   = replace_na(future_mw, 0)
  )

## Optionally, country outlines could be added for context, but we omit them here
## to keep the figure focused on basins and biomes.
```

```{r plot-basin-biome}
#| warning: false
#| message: false
ggplot() +
  # Biome latitude bands
  geom_rect(
    data = lat_bands,
    aes(xmin = -180, xmax = 180, ymin = ymin, ymax = ymax, fill = biome),
    color = NA,
    alpha = 0.35
  ) +
  scale_fill_manual(values = pal_ci, name = "Biome (Li & He)") +
  new_scale_fill() +
  # Basins shaded by total future capacity (MW)
  geom_sf(
    data = basin_map,
    aes(fill = future_mw),
    color = NA,
    inherit.aes = FALSE
  ) +
  scale_fill_gradientn(
    colours = pal_capacity,
    trans = "sqrt",
    na.value = "grey95",
    labels = comma,
    name = "Future capacity\n(MW)"
  ) +
  # Future dams as points — all projects colored by basin (no explicit top-50 emphasis)
  geom_point(
    data = future_geo,
    aes(
      x = Lon_Cleaned,
      y = LAT_cleaned,
      size = `Capacity (MW)`,
      color = basin_group
    ),
    shape = 16,
    alpha = 0.85,
    inherit.aes = FALSE
  ) +
  scale_size_continuous(
    range = c(1.2, 9),
    labels = comma,
    name = "Dam capacity (MW)"
  ) +
  scale_color_manual(
    values = pal_basin10,
    guide = "none"
  ) +
  # Labels for top 10 basins with repelled positions and leader lines
  ggrepel::geom_text_repel(
    data = basin_labels,
    aes(x = anchor_x, y = anchor_y, label = label),
    color = "#000000",
    size = 3,
    fontface = "bold",
    family = "Times New Roman",
    lineheight = 0.9,
    seed = 123,
    nudge_x = basin_labels$nudge_x,
    nudge_y = basin_labels$nudge_y,
    box.padding = 0.5,
    point.padding = 0.5,
    segment.color = "#000000",
    segment.size = 0.4,
    segment.alpha = 0.95,
    segment.curvature = 0.2,
    segment.angle = 20,
    segment.ncp = 3,
    min.segment.length = 0,
    max.overlaps = 20,
    inherit.aes = FALSE
  ) +
  coord_sf(xlim = c(-180, 180), ylim = c(-60, 75)) +
  labs(
    title = "Future hydropower: basins, biomes, and countries on one map",
    subtitle = "Background bands = biome (Li & He). Basins (BasinATLAS lev02) = total future capacity (MW; FHReD). Points = dams colored by FHReD major basin; size = capacity (MW).",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )
```

**Takeaway:** On one map you see **biomes** (latitude bands), **basins** (BasinATLAS lev02 polygons shaded by total future capacity, MW), and **projects** (dam points colored by top 10 river basins). We omit the high-latitude **tundra** band here because very few future dams fall in that biome, and the final figure will zoom into lower latitudes where most capacity is planned. Most planned hydropower is in a handful of megabasins in tropical and temperate biomes, where Li & He (2022) report higher carbon intensity.

